### 2024-05-29 manuscript2.Rmd
29/5

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: "word_document"
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

## Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study
### Background
Trauma is a leading cause of morbidity and mortality worldwide. A cornerstone in trauma quality improvement programs is multidisciplinary morbidity and mortality conferences. The purpose of these conferences is to identify opportunities for improvement (OFI). Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about OFI in this group of patients. The aim of this study was to characterize OFI in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with OFI in these patients.

### Methods
A registry-based study of trauma patients admitted to the ICU between 2014-2023 at the Karolinska University Hospital in Solna, Sweden, who were reviewed regarding the presence of OFI in a morbidity and mortality conference. Bivariable and multivariable logistic regression was used to determine associations between selected patient and process factors and OFI.

### Results
OFI was identified in 143 (9,9%) out of 1449 patients. Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with increased odds of OFI.

### Conclusion
Several patient and process factors were found to be associated with OFI, indicating that patients who were relatively more stable upon arrival, patients with prolonged ICU stay, patients with prolonged mechanical ventilation, and patients who were alive 30 days after injury had higher odds of OFI.

# Abbreviations

ASA score = American Society of Anesthesiologists score
CT = computed tomography
GCS = Glasgow Coma Scale
ISS = Injury Severity Score
OFI = Opportunity For Improvement
RR = respiratory rate
RTS = Revised Trauma Score
SBP = systolic blood pressure
SweTrau = Swedish Trauma Registry
TQIP = Trauma Quality Improvement Programs

# Introduction

Trauma — the clinical entity of injury and the body’s associated response — is a leading cause of morbidity and mortality worldwide [@60] and is one of the top contributors to disease burden worldwide, measured by disability-adjusted life-years (DALYs) [@6]. Injury-related deaths constitute 7.6% of all deaths and take the lives of 4.3 million people yearly [@52, @1].

The paradigm of Donabedian states that implementing proven structures and processes of care improves patient outcomes [@8]. Trauma Quality Improvement Programs (TQIP) emerged in high-income countries to enhance the quality of in-hospital trauma care and aim to identify areas to implement such structures and processes of care [@47]. There are several TQIP interventions, out of which morbidity and mortality review is a key method of addressing all necessary components of trauma improvement [@10]. The morbidity and mortality review aims to assess preventable deaths and identify opportunities for improvement (OFI) in trauma care. Common OFIs among trauma patients relates to airway and ventilator management, fluid resuscitation, delays to treatment and error in judgement [@53, @67, @68].

Trauma patients are often admitted to the intensive care unit (ICU) due to the severity of their injuries and the subsequent physiological and anatomical sequelae [@64]. Prior research has focused on classifying errors in preventable trauma deaths, with clinical performance errors, often stemming from a lack of knowledge, being identified as the most common type, particularly in the emergency department [@62]. However, reporting of errors in trauma resuscitation varies between healthcare centers [@63]. As a results, there is a notable gap in research regarding error reporting practices in trauma care — especially for trauma patients admitted to the ICU [@12, @64]. Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their complex medical backgrounds as well as the complexity and duration of  care, making it difficult to establish the standard of care for each individual. Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9]. While numerous studies have explored various facets of trauma care enhancement, research specifically centered on OFIs in ICU-admitted trauma patients remains limited [@64].

The aim of this study was to characterize OFI in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with OFI in these patients.

# Methods

## Study design

We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which reports to the Swedish Trauma registry (SweTrau), and to the hospital's local trauma care quality database. From 2017 the trauma care quality database includes all patients in the Swedish trauma registry as well as the results of the morbidity and mortality review, including presence or absence of OFI.

## Setting

The Karolinska University Hospital in Solna is a level one equivalent trauma center designated to receive all severely injured patients in the wider metropolitan area of Stockholm [@17]. The hospital has direct access to radiology, surgery, emergency medicine, intensive care, and consultants from relevant specialities [@18]. All patients are included in a morbidity and mortality screening process, combining audit filters and an individual review by specialized nurses. Those identified as having a higher likelihood of OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for improved organization or greater involvement of senior staff in trauma care. The presence or absence of OFI is decided among the participants of the conference and subsequently documented in the trauma care quality database.

## Participants

The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU. We excluded patients who were younger than 15 years and patients who were dead on arrival. Patients under 15 years of age were not included in the study as they are included in other registries, and because the clinical approach for children differs from that of adults.

## Variables

### Study outcome

The study’s outcome was the presence of OFI — defined as “Yes” for at least one identified OFI or “No” for no identified OFI — based on the consensus reached during the morbidity and mortality conference. An OFI is a failure of care, such as a technical error, clinical judgement error or any preventable death.

### Patient and process factors

The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, and factors registered in SweTrau. Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality. Sex was categorized as male sex or female sex. Mechanical ventilation was stratified into “No intubation”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”. On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday. ICU length of stay was categorized as “≤ 7 days” or “> 7 days”. ASA preinjury was divided into “ASA 1-2” and “ASA 3-6”. 30-day mortality was determined based on whether the patient was deceased or alive 30 days after injury.

The continuous variables in the analysis were age, Revised Trauma Score (RTS), ISS, and time to first computed tomography (CT). RTS is a measure of severity based on prehospital or emergency department values. In instances where hospital values for systolic blood pressure (SBP), respiratory rate (RR), or Glasgow Come Scale (GCS) scores were missing from the emergency department records, prehospital values were used instead. If the patient was intubated, it was assumed that their GCS was 3 and RR was 0. Subsequently, for the RTS variable, each of the variables GCS, SBP, and RR were converted to RTS codes ranging from 0 to 4 based on their respective values. RTS was then calculated using the formula for RTS: 0.9368 * GCS-code + 0.7326 SBP-code + 0.2908 RR-code to derive the RTS value.

## Statistical analysis

When conducting logistic regression analysis, the number of events per variable (EPV) equal to 10 is a rule of thumb. If the EPV falls below this threshold, the results for regression analyses should be interpreted with caution [@45]. Keeping this guideline in mind, we selected independent variables for the analysis. To achieve this, we converted RTS, ISS, and time to first CT into continuous variables, while categorical variables were divided into as few levels as possible.

A complete case analysis was conducted after handling missing values for SBP, RR and GCS as described above. We presented sample characteristics using descriptive statistics — mean for continuous variables and prevalence for categorical variables. Pearson’s Chi-squared test was used to calculate p-values for categorical variables, while the Wilcoxon rank sum test was used for continuous variables. Continuous variables were compared based on their mean values and standard deviations, while categorical variables were assessed by the number of patients within each subgroup and the corresponding subgroup percentages.

Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI. Odds ratios (OR) with associated 95% confidence intervals were used with a 5% significance level. For continuous variables, ORs were derived by evaluating each unitary step increment. A subgroup analysis of patients who were alive 30 days after the injury was conducted to reduce impact from death as a cause of the association between OFI and patient and process factors. All statistical analyses were conducted using R [@19]. The analyses were initially conducted on synthetic data and subsequently applied to data from SweTrau and the trauma care quality database to ensure objectivity.

# Results

## Participants
Figure 1. Flowchart describing the inclusion and exclusion criteria of the study

```{r, include=FALSE}
source("functions/flowchart.R")
```

```{r, echo=FALSE}
for (i in 1:(length(vert) - 1)) {
  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
    print
}

connectGrob(vert$eligible, excluded, type = "L")
connectGrob(vert$included, excluded1, type = "L")

vert
excluded
excluded1
```

The patient demographics of this study is presented in Figure 1. A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded. Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of OFI. This left us with a total of 1,742 patients eligible for the study. Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study.

## Descriptive data

```{r include_scripts, include=FALSE}
source("functions/table1.R")
```

```{r, echo=FALSE}
table2
```

Table 1 present sample characteristics. Of all patients, males constituted 1,096 (76%) and females constituted 353 (24%) patients. There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than seven days. The number of patients who arrived outside working hours was 1,042 (72%). A total of 1,064 (73%) patients were equal to or less than seven days in the ICU, and 385 (27%) patients were more than seven days in the ICU. A total of 1,080 (75%) patients had an ASA score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA score of 3-6 before arrival to the hospital. Amongst our patients, 30-day mortality was at 286 (20%).

When comparing patients with OFI to those without, significant differences emerged across various factors. Specifically, individuals with OFI were less likely to require intubation, had higher RTS scores, experienced shorter time to initial CT scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.

## Main results

```{r, include=FALSE}
source("functions/table2.R")
```

```{r, echo=FALSE}
table3_merge
```

Table 2 shows the unadjusted and adjusted associations of the selected patient and process factors in relation to OFI. In the unadjusted analysis, the factors that were significantly associated with the presence of OFI were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.

In the adjusted analysis the following factors demonstrated significant association with increasing odds of OFI: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.

Sex, age, ISS, time to first CT, on-call hours, and ASA preinjury were not significantly associated with OFI in the unadjusted nor the adjusted analyses.

The same patient and process factors were significant in both analyses. The ORs increased for all factors in the adjusted analysis compared to the unadjusted analysis except for survival 30 days after injury where there was a slight decrease.

```{r, include=FALSE}
source("functions/table3.R")
```

```{r, echo=FALSE}
table3alive_merge
```

Table 3 shows the unadjusted and adjusted associations of OFI and the selected patient and process factors among the patients who were alive 30 days after injury. In the unadjusted analysis > 7 days in the ICU was significantly associated with OFI. In the adjusted analysis increased RTS-score was significantly associated with increased odds of OFI.

# Discussion

The key results from this study are that higher RTS, > 7 days length of stay in the ICU and survival 30 days after injury were significantly associated with higher odds of OFI. Age, sex, mechanical ventilation, ISS, time to first CT, arrival during on-call hours, and ASA preinjury were not significantly associated with OFI in either the adjusted or the unadjusted analyses. In the subgroup analysis of patients who were alive 30 days after injury, no patient or process factor was significantly associated with OFI in the adjusted analysis.

We found that higher RTS was associated with increased odds of OFI. Previous research has shown that patients with relatively better vital parameter values, had a significant association with increased odds of OFI in the unadjusted analysis, but not in the adjusted analysis [@12]. These findings could be because patients with lower RTS are often in critical condition upon arrival — upon which emergency interventions are performed immediately. Conversely, patients with higher RTS, indicating relative stability in this patient group upon arrival, may experience delayed interventions.

Similarly, prolonged ICU stay was found to be significantly associated with OFI. There are two things to consider regarding this variable. Firstly, patients requiring extended ICU care are often those with complex medical conditions, severe injuries, or critical illness. As a result, their prolonged stay in the ICU may signify a higher baseline risk for experiencing complications, increasing odds of OFI. Additionally, prolonged ICU stay increases likelihood of medical errors and adverse events to occur. The ICU environment, while designed to provide advanced medical care, can also be fraught with complexities, including high patient acuity, numerous interventions, and a fast-paced workflow. The longer a patient remains in the ICU, the greater the cumulative exposure to potential errors. Conversely, shorter ICU stays may indicate patient demise before potential mistakes could occur, particularly in cases where mortality was anticipated rather than preventable. This finding is similar to previous research demonstrating prolonged ICU stays as a risk factor for increased morbidity and mortality in trauma patients [@22, @26]. Furthermore, it also corresponds with a retrospective cohort indicating that lower initial illness severity and younger age are associated with earlier discharge, further supporting the link between higher odds of OFI and prolonged ICU stays [@25].

The reason for prolonged ICU stays being associated with increased mortality has been investigated by previous studies suggesting that factors such as comorbidities, age, and illness severity upon admission do not fully elucidate the increased long-term mortality observed in this patient cohort. As the duration of ICU stay increases, the primary reason for ICU stay becomes less linked to the patient’s original condition and increasingly tied to persistent chronic illness accompanied by subsequent organ dysfunction. The cause of the organ dysfunction comes from increased exposure to iatrogenous factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions. [@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and ISS in our study These predictors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and ISS.

Another factor associated with increased odds of OFI was survival 30 days after injury. Previous research has identified an association between increased mortality and the severity of the injury upon admission [@59], suggesting that patients at high risk of 30-day mortality are typically in poor health upon arrival. Consequently, the likelihood of OFI decreases as deaths may not be preventable.

We also found that mechanical ventilation more than seven days was significantly associated with increased odds of OFI. Previous research has shown that prolonged mechanical ventilation is associated with increased mortality. This is because mechanical ventilation is associated with severe complications and failure of weaning further increases length of mechanical ventilation as well as ICU length of stay [@65, @66]. Our finding of this association may thus be because of the increased rate of complications and mortality associated with prolonged mechanical ventilation.

Several limitations need to be acknowledged in our study. This was a single-center study and primarily provide insights into the context of the Karolinska University Hospital in Solna. Another limitation is the selection of patients for the morbidity and mortality review, which heavily relied on the local audit filters. The use of audit filters for patient selection for the morbidity and mortality review is associated with high false positive rates, ranging from 24% to 80% [@38]. Even though the use of audit filters is associated with high false positive rates, there is still a risk of misplacement where some patients with OFI may have been overlooked. Furthermore, the handling of the variables to the RTS score – namely RR, SBP, and GCS – introduced potential sources of variability. The use of prehospital, emergency department, or intubation-based values for these variables may have affected data accuracy and reproducibility.

# Conclusion
The findings of this study revealed a significant association between higher RTS scores, ICU stay exceeding seven days, mechanical ventilation exceeding seven days, and survival 30 days after injury with increased odds of OFI. These factors represent viable targets for new initiatives to further improve the quality of trauma care.

# References

### 2024-06-03 manuscript2.Rmd
3/6

@@
 title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
-output: "word_document"
+output: "html_document"
 author: "Elin Sun Cao"
@@
 Trauma is a leading cause of morbidity and mortality worldwide.
-A cornerstone in trauma quality improvement programs is multidisciplinary morbidity and mortality conferences.
-The purpose of these conferences is to identify opportunities for improvement (OFI).
-Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about OFI in this group of patients.
-The aim of this study was to characterize OFI in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with OFI in these patients.
+A cornerstone in trauma quality improvement programs is multidisciplinary morbidity and mortality review The purpose of these reviews is to identify opportunities for improvement (opportunities for improvement) and corresponding corrective actions.
+Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this cohort.
+The aim of this study was to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
 
 ### Methods
-A registry-based study of trauma patients admitted to the ICU between 2014-2023 at the Karolinska University Hospital in Solna, Sweden, who were reviewed regarding the presence of OFI in a morbidity and mortality conference.
-Bivariable and multivariable logistic regression was used to determine associations between selected patient and process factors and OFI.
+A registry-based study of trauma patients admitted to the ICU between 2014-2023 at the Karolinska University Hospital in Solna, Sweden, who were reviewed regarding the presence of opportunities for improvement in a morbidity and mortality review.
+Univariate and multivariable logistic regression was used to determine associations between selected patient and process factors and opportunities for improvement.
 
 ### Results
-OFI was identified in 143 (9,9%) out of 1449 patients.
-Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with increased odds of OFI.
+opportunities for improvement was identified in 143 (9,9%) out of 1449 patients.
+Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with increased odds of opportunities for improvement.
 
 ### Conclusion
-Several patient and process factors were found to be associated with OFI, indicating that patients who were relatively more stable upon arrival, patients with prolonged ICU stay, patients with prolonged mechanical ventilation, and patients who were alive 30 days after injury had higher odds of OFI.
-
-# Abbreviations
-
-ASA score = American Society of Anesthesiologists score
-CT = computed tomography
-GCS = Glasgow Coma Scale
-ISS = Injury Severity Score
-OFI = Opportunity For Improvement
-RR = respiratory rate
-RTS = Revised Trauma Score
-SBP = systolic blood pressure
-SweTrau = Swedish Trauma Registry
-TQIP = Trauma Quality Improvement Programs
+Several patient and process factors were found to be associated with opportunities for improvement, indicating that patients who were relatively more stable upon arrival, patients with prolonged ICU stay, patients with prolonged mechanical ventilation, and patients who were alive 30 days after injury had higher odds of opportunities for improvement.
 
@@
 
-Trauma — the clinical entity of injury and the body’s associated response — is a leading cause of morbidity and mortality worldwide [@60] and is one of the top contributors to disease burden worldwide, measured by disability-adjusted life-years (DALYs) [@6].
-Injury-related deaths constitute 7.6% of all deaths and take the lives of 4.3 million people yearly [@52, @1].
+Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of morbidity and mortality worldwide [@60, @6].
+Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and result in complications or death [@64, @69].
+While these complications are often expected, the complexity of ICU-care requires providers to be highly adaptable and makes it prone to mistakes and errors.
+Peer review is a cornerstone of trauma quality improvement programs to identify preventable causes of morbidity and mortality in trauma patient [@70], referred to as opportunities for improvement [*ref* - 10 i texten].
+Previous research on opportunities for improvement in trauma care has often focused on the initial management and have identified airway and ventilator management, fluid resuscitation, delays to treatment and judgement errors as common areas for opportunities for improvement [*ref 10*].
 
-The paradigm of Donabedian states that implementing proven structures and processes of care improves patient outcomes [@8].
-Trauma Quality Improvement Programs (TQIP) emerged in high-income countries to enhance the quality of in-hospital trauma care and aim to identify areas to implement such structures and processes of care [@47].
-There are several TQIP interventions, out of which morbidity and mortality review is a key method of addressing all necessary components of trauma improvement [@10].
-The morbidity and mortality review aims to assess preventable deaths and identify opportunities for improvement (OFI) in trauma care.
-Common OFIs among trauma patients relates to airway and ventilator management, fluid resuscitation, delays to treatment and error in judgement [@53, @67, @68].
-
-Trauma patients are often admitted to the intensive care unit (ICU) due to the severity of their injuries and the subsequent physiological and anatomical sequelae [@64].
 Prior research has focused on classifying errors in preventable trauma deaths, with clinical performance errors, often stemming from a lack of knowledge, being identified as the most common type, particularly in the emergency department [@62].
-However, reporting of errors in trauma resuscitation varies between healthcare centers [@63].
+However, reporting of errors in trauma resuscitation varies between settings and studies [@63].
 As a results, there is a notable gap in research regarding error reporting practices in trauma care — especially for trauma patients admitted to the ICU [@12, @64].
-Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their complex medical backgrounds as well as the complexity and duration of  care, making it difficult to establish the standard of care for each individual.
+Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their medical complexity, making it difficult to establish the standard of care for each individual.
 Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9].
-While numerous studies have explored various facets of trauma care enhancement, research specifically centered on OFIs in ICU-admitted trauma patients remains limited [@64].
-
-The aim of this study was to characterize OFI in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with OFI in these patients.
+While numerous studies have explored various facets of trauma care enhancement, research specifically centered on opportunities for improvement in ICU-admitted trauma patients remains limited [@64].
+The aim of this study was to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
 
@@
 
-We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which reports to the Swedish Trauma registry (SweTrau), and to the hospital's local trauma care quality database.
-From 2017 the trauma care quality database includes all patients in the Swedish trauma registry as well as the results of the morbidity and mortality review, including presence or absence of OFI.
+We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which is included in the Swedish Trauma registry (SweTrau), and the hospital's local trauma care quality database.
 
@@
 
-The Karolinska University Hospital in Solna is a level one equivalent trauma center designated to receive all severely injured patients in the wider metropolitan area of Stockholm [@17].
-The hospital has direct access to radiology, surgery, emergency medicine, intensive care, and consultants from relevant specialities [@18].
-All patients are included in a morbidity and mortality screening process, combining audit filters and an individual review by specialized nurses.
-Those identified as having a higher likelihood of OFI are discussed in a multidisciplinary conference held every 6-8 weeks.
-Examples of OFI include the need for improved organization or greater involvement of senior staff in trauma care.
-The presence or absence of OFI is decided among the participants of the conference and subsequently documented in the trauma care quality database.
+The Karolinska University Hospital in Solna is the trauma center designated to receive all severely injured patients in the wider metropolitan area of Stockholm [@17].
+The hospital has direct access to radiology, surgery, intensive care, and consultants from relevant specialities [@18].
+The care of adult trauma patients is screened for opportunities for improvement using a combination of audit filters and individual review by specialized nurses.
+Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review held every 6-8 weeks.
+The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database.
+Before 2017 a non-systematic selection from clinicians involved in the opportunities for improvement process was done by filtering for patients that were thought to have a high risk for opportunities for improvement, but starting in 2017 the care of all patients is screened.
 
@@
 The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9.
-We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU.
+We included all patients who had been included in the opportunities for improvement screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU.
 We excluded patients who were younger than 15 years and patients who were dead on arrival.
-Patients under 15 years of age were not included in the study as they are included in other registries, and because the clinical approach for children differs from that of adults.
+Patients under 15 years of age were not included in the study as their clinical and review pathway differ compared to adults.
+
+## Study size
+
+When conducting logistic regression analysis, the number of events per variable (EPV) equal to 10 is a rule of thumb.
+If the EPV falls below this threshold, the results for regression analyses should be interpreted with caution [@45].
 
@@
 
+Keeping this guideline in mind, we selected independent variables for the analysis.
+To achieve this, we converted RTS, ISS, and time to first CT into continuous variables, while categorical variables were divided into as few levels as possible.
+
 ### Study outcome
 
-The study’s outcome was the presence of OFI — defined as “Yes” for at least one identified OFI or “No” for no identified OFI — based on the consensus reached during the morbidity and mortality conference.
-An OFI is a failure of care, such as a technical error, clinical judgement error or any preventable death.
+The study’s outcome was the presence of any opportunity for improvement.
+An opportunity for improvement presents preventable errors in care with direct adverse effects on patient outcomes or recurrent deviations from safe clinical practice.
+Examples include technical error, clinical judgement error or any preventable death.
 
@@
 
-The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, and factors registered in SweTrau.
-Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality.
+The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion.Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality.
 Sex was categorized as male sex or female sex.
-Mechanical ventilation was stratified into “No intubation”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”.
+Mechanical ventilation was stratified into “Not intubated”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”.
 On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday.
@@
 
-The continuous variables in the analysis were age, Revised Trauma Score (RTS), ISS, and time to first computed tomography (CT).
+The continuous variables in the analysis were age in years, Revised Trauma Score (RTS), ISS, and time to first computed tomography in minutes.
 RTS is a measure of severity based on prehospital or emergency department values.
@@
 If the patient was intubated, it was assumed that their GCS was 3 and RR was 0.
-Subsequently, for the RTS variable, each of the variables GCS, SBP, and RR were converted to RTS codes ranging from 0 to 4 based on their respective values.
-RTS was then calculated using the formula for RTS: 0.9368 * GCS-code + 0.7326 SBP-code + 0.2908 RR-code to derive the RTS value.
+Subsequently, for the RTS variable, each of the variables GCS, SBP, and RR were converted to RTS codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].
 
@@
 
-When conducting logistic regression analysis, the number of events per variable (EPV) equal to 10 is a rule of thumb.
-If the EPV falls below this threshold, the results for regression analyses should be interpreted with caution [@45].
-Keeping this guideline in mind, we selected independent variables for the analysis.
-To achieve this, we converted RTS, ISS, and time to first CT into continuous variables, while categorical variables were divided into as few levels as possible.
+We conducted complete case analysis after handling missing values for SBP, RR and GCS as described above.
+Pearson’s Chi-squared test and the Wilcoxon rank sum test were used to calculate the significance for categorical and continuous variables respectively.
+Continuous variables are presented using mean values and standard deviations,  and categorical variables are presented using the number of patients within each subgroup and the corresponding subgroup percentages.
 
-A complete case analysis was conducted after handling missing values for SBP, RR and GCS as described above.
-We presented sample characteristics using descriptive statistics — mean for continuous variables and prevalence for categorical variables.
-Pearson’s Chi-squared test was used to calculate p-values for categorical variables, while the Wilcoxon rank sum test was used for continuous variables.
-Continuous variables were compared based on their mean values and standard deviations, while categorical variables were assessed by the number of patients within each subgroup and the corresponding subgroup percentages.
-
-Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
-Odds ratios (OR) with associated 95% confidence intervals were used with a 5% significance level.
+Univariate logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and opportunities for improvement.
+Odds ratios (OR) with associated 95% confidence intervals were usecalculated.
+A p-value of less than 0.05 considered indicative of a significant difference.
 For continuous variables, ORs were derived by evaluating each unitary step increment.
-A subgroup analysis of patients who were alive 30 days after the injury was conducted to reduce impact from death as a cause of the association between OFI and patient and process factors.
-All statistical analyses were conducted using R [@19].
-The analyses were initially conducted on synthetic data and subsequently applied to data from SweTrau and the trauma care quality database to ensure objectivity.
+A subgroup analysis of patients who were alive 30 days after the injury was conducted to reduce impact from death as a cause of the association between opportunities for improvement and patient and process factors.
+All statistical analyses were conducted using R version 4.4 [@19].
 
@@
 A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded.
-Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of OFI.
-This left us with a total of 1,742 patients eligible for the study.
+Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement This left us with a total of 1,742 patients eligible for the study.
 Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study.
@@
 Table 1 present sample characteristics.
-Of all patients, males constituted 1,096 (76%) and females constituted 353 (24%) patients.
+Most patients were male 1,096 (76%) vs 353 (24%) female patients.
 There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than seven days.
@@
 
-When comparing patients with OFI to those without, significant differences emerged across various factors.
-Specifically, individuals with OFI were less likely to require intubation, had higher RTS scores, experienced shorter time to initial CT scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
+When comparing patients with opportunities for improvement to those without, differences emerged across various factors.
+Specifically, individuals with opportunities for improvement were less likely to require intubation, had higher RTS scores, experienced shorter time to initial CT scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
 
@@
 
-Table 2 shows the unadjusted and adjusted associations of the selected patient and process factors in relation to OFI.
-In the unadjusted analysis, the factors that were significantly associated with the presence of OFI were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.
+Table 2 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement.
+In the unadjusted analysis, the factors that were significantly associated with the presence of opportunities for improvement were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.
 
-In the adjusted analysis the following factors demonstrated significant association with increasing odds of OFI: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
+In the adjusted analysis the following factors demonstrated significant association with increasing odds of opportunities for improvement: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
 
-Sex, age, ISS, time to first CT, on-call hours, and ASA preinjury were not significantly associated with OFI in the unadjusted nor the adjusted analyses.
-
-The same patient and process factors were significant in both analyses.
-The ORs increased for all factors in the adjusted analysis compared to the unadjusted analysis except for survival 30 days after injury where there was a slight decrease.
+Sex, age, ISS, time to first CT, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
 
@@
 
-Table 3 shows the unadjusted and adjusted associations of OFI and the selected patient and process factors among the patients who were alive 30 days after injury.
-In the unadjusted analysis > 7 days in the ICU was significantly associated with OFI.
-In the adjusted analysis increased RTS-score was significantly associated with increased odds of OFI.
+Table 3 shows the unadjusted and adjusted associations of opportunities for improvement and the selected patient and process factors among the patients who were alive 30 days after injury.
+In the unadjusted analysis > 7 days in the ICU was significantly associated with opportunities for improvement.
+In the adjusted analysis increased RTS-score was significantly associated with increased odds of opportunities for improvement.
 
@@
 
-The key results from this study are that higher RTS, > 7 days length of stay in the ICU and survival 30 days after injury were significantly associated with higher odds of OFI.
-Age, sex, mechanical ventilation, ISS, time to first CT, arrival during on-call hours, and ASA preinjury were not significantly associated with OFI in either the adjusted or the unadjusted analyses.
-In the subgroup analysis of patients who were alive 30 days after injury, no patient or process factor was significantly associated with OFI in the adjusted analysis.
+The key results from this study are that higher RTS, > 7 days length of stay in the ICU and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
+Age, sex, mechanical ventilation, ISS, time to first CT, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in either the adjusted or the unadjusted analyses.
+In the subgroup analysis of patients who were alive 30 days after injury, no patient or process factor was significantly associated with opportunities for improvement in the adjusted analysis.
 
-We found that higher RTS was associated with increased odds of OFI.
-Previous research has shown that patients with relatively better vital parameter values, had a significant association with increased odds of OFI in the unadjusted analysis, but not in the adjusted analysis [@12].
+We found that higher RTS was associated with increased odds of opportunities for improvement.
+Previous research has shown that patients with relatively better vital parameter values, had a significant association with increased odds of opportunities for improvement in the unadjusted analysis, but not in the adjusted analysis [@12].
 These findings could be because patients with lower RTS are often in critical condition upon arrival — upon which emergency interventions are performed immediately.
@@
 
-Similarly, prolonged ICU stay was found to be significantly associated with OFI.
+Similarly, prolonged ICU stay was found to be significantly associated with opportunities for improvement.
 There are two things to consider regarding this variable.
 Firstly, patients requiring extended ICU care are often those with complex medical conditions, severe injuries, or critical illness.
-As a result, their prolonged stay in the ICU may signify a higher baseline risk for experiencing complications, increasing odds of OFI.
-Additionally, prolonged ICU stay increases likelihood of medical errors and adverse events to occur.
+As a result, their prolonged stay in the ICU may signify a higher baseline risk for experiencing complications, increasing odds of opportunities for improvement.
+Additionally, prolonged ICU stay increases likelihood of medical errors and adverse events to occur [*REF*].
 The ICU environment, while designed to provide advanced medical care, can also be fraught with complexities, including high patient acuity, numerous interventions, and a fast-paced workflow.
@@
 This finding is similar to previous research demonstrating prolonged ICU stays as a risk factor for increased morbidity and mortality in trauma patients [@22, @26].
-Furthermore, it also corresponds with a retrospective cohort indicating that lower initial illness severity and younger age are associated with earlier discharge, further supporting the link between higher odds of OFI and prolonged ICU stays [@25].
+Furthermore, it also corresponds with a retrospective cohort indicating that lower initial illness severity and younger age are associated with earlier discharge, further supporting the link between higher odds of opportunities for improvement and prolonged ICU stays [@25].
 
@@
 As the duration of ICU stay increases, the primary reason for ICU stay becomes less linked to the patient’s original condition and increasingly tied to persistent chronic illness accompanied by subsequent organ dysfunction.
-The cause of the organ dysfunction comes from increased exposure to iatrogenous factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions.
-[@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and ISS in our study These predictors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and ISS.
+The cause of the organ dysfunction comes from increased exposure to iatrogenic factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions.
+[@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and ISS in our study.
+These factors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and ISS.
 
-Another factor associated with increased odds of OFI was survival 30 days after injury.
+Another factor associated with increased odds of opportunities for improvement was survival 30 days after injury.
 Previous research has identified an association between increased mortality and the severity of the injury upon admission [@59], suggesting that patients at high risk of 30-day mortality are typically in poor health upon arrival.
-Consequently, the likelihood of OFI decreases as deaths may not be preventable.
+Consequently, the likelihood of opportunities for improvement decreases as deaths may not be preventable.
 
-We also found that mechanical ventilation more than seven days was significantly associated with increased odds of OFI.
+We also found that mechanical ventilation more than seven days was significantly associated with decreased odds of opportunities for improvement.
 Previous research has shown that prolonged mechanical ventilation is associated with increased mortality.
@@
 Our finding of this association may thus be because of the increased rate of complications and mortality associated with prolonged mechanical ventilation.
+*skriv om så det gäller minskad OFI*
 
@@
 The use of audit filters for patient selection for the morbidity and mortality review is associated with high false positive rates, ranging from 24% to 80% [@38].
-Even though the use of audit filters is associated with high false positive rates, there is still a risk of misplacement where some patients with OFI may have been overlooked.
+Even though the use of audit filters is associated with high false positive rates, there is still a risk of misplacement where some patients with opportunities for improvement may have been overlooked.
 Furthermore, the handling of the variables to the RTS score – namely RR, SBP, and GCS – introduced potential sources of variability.
@@
 # Conclusion
-The findings of this study revealed a significant association between higher RTS scores, ICU stay exceeding seven days, mechanical ventilation exceeding seven days, and survival 30 days after injury with increased odds of OFI.
+The findings of this study revealed a significant association between higher RTS scores, ICU stay exceeding seven days, mechanical ventilation exceeding seven days, and survival 30 days after injury with increased odds of opportunities for improvement.
 These factors represent viable targets for new initiatives to further improve the quality of trauma care.

### 2024-06-03 functions/flowchart.R

@@ -55,15 +55,18 @@ excluded1 <- moveBox(excluded1,
                      x = 0.8,
                      y = 0.4)
 
+#Defining small arrow
+small_arrow <- arrow(length = unit(0.1, "inches"), type = "closed")
+
 # Connect boxes vertically
 for (i in 1:(length(vert) - 1)) {
-  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert", arrow = small_arrow) %>%
     print
 }
 
 # Connect excluded box horizontally
-connectGrob(vert$eligible, excluded, type = "L")
-connectGrob(vert$included, excluded1, type = "L")
+connectGrob(vert$eligible, excluded, type = "L", arrow = small_arrow)
+connectGrob(vert$included, excluded1, type = "L", arrow = small_arrow)
 
 # Print boxes
 vert

### 2024-06-03 functions/table1.R

@@ -210,8 +210,8 @@ table2 <- table1 %>%
   ) %>%
   bold_labels() %>% 
   add_overall(last = TRUE) %>% 
-  add_p() %>%
-  bold_p(t=0.05) %>%
+  #add_p() %>%
+  #bold_p(t=0.05) %>%
   modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 1. Sample Characteristics</div>") %>% 
  # as_flex_table() %>%
   print()

### 2024-06-03 functions/table2.R

@@ -26,7 +26,8 @@ table3a <- tbl_uvregression(data = tablereg,
                               TimeFCT = "Time to first CT"
                             )) %>%
   bold_labels() %>%
-  bold_p(t = 0.05) 
+  bold_p(t = 0.05) %>%
+  hide_n = TRUE
 
 #print(table3a)
 
@@ -43,14 +44,16 @@ table3b <- tbl_regression(adjusted_table,
                                        ASApreinjury = "ASA preinjury",
                                        TimeFCT = "Time to first CT")) %>%
   bold_labels() %>%
-  bold_p(t = 0.05)
+  bold_p(t = 0.05) %>%
+  hide_n = TRUE
 
 # print(table3b)
 
 # Merging Tables
 table3_merge <- tbl_merge(tbls = list(table3a, table3b),
                           tab_spanner = c("**Unadjusted**", "**Adjusted**")) %>%
-  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 2. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement</div>")
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 2. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement (N = 1200).</div>")
+
 
 print(table3_merge)
 

### 2024-06-03 functions/table3.R

@@ -9,28 +9,29 @@ tablereg1 <- ofi_alive %>%
 tablereg1$daysinICU <- fct_relevel(tablereg1$daysinICU, "≤ 7 days", "> 7 days")
 
 # Unadjusted Table
-table3aalive <- tbl_uvregression(data = tablereg1,
-                            method = glm,
-                            y = OpportunityForImprovement1,
-                            method.args = list(family = binomial),
-                            exponentiate = TRUE, 
-                            label = list(
-                              RTS = "Revised Trauma Score",
-                              daysinICU = "Days in the ICU",
-                              TimeFInt = "Time to first intervention",
-                              ASApreinjury = "ASA preinjury",
-                              OnDuty = "On call hours",
-                              Intubation = "Mechanical ventilation",
-                              TimeFCT = "Time to first CT"
-                            )) %>%
-  bold_labels() %>%
-  bold_p(t = 0.05) 
+#table3aalive <- tbl_uvregression(data = tablereg1,
+#                            method = glm,
+#                            y = OpportunityForImprovement1,
+#                            method.args = list(family = binomial),
+#                            exponentiate = TRUE, 
+#                            label = list(
+#                              RTS = "Revised Trauma Score",
+#                              daysinICU = "Days in the ICU",
+#                              TimeFInt = "Time to first intervention",
+#                              ASApreinjury = "ASA preinjury",
+#                              OnDuty = "On call hours",
+#                              Intubation = "Mechanical ventilation",
+#                              TimeFCT = "Time to first CT"
+#                            )) %>%
+#  bold_labels() %>%
+#  bold_p(t = 0.05) %>%
+#  hide_n = TRUE
 
 # Adjusted Table
 #Creating linear regression 
 adjusted_table1 <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury, family = binomial, data = tablereg1) 
 
-table3balive <- tbl_regression(adjusted_table1,
+table3alive <- tbl_regression(adjusted_table1,
                           exponentiate = TRUE, 
                           label = list(RTS = "Revised Trauma Score",
                                        daysinICU = "Days in the ICU",
@@ -39,12 +40,37 @@ table3balive <- tbl_regression(adjusted_table1,
[…]

### 2024-06-05 manuscript2.Rmd
5/6

@@
 title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
-output: "html_document"
+output: "word_document"
 author: "Elin Sun Cao"
@@
 ### Background
-Trauma is a leading cause of morbidity and mortality worldwide.
-A cornerstone in trauma quality improvement programs is multidisciplinary morbidity and mortality review The purpose of these reviews is to identify opportunities for improvement (opportunities for improvement) and corresponding corrective actions.
+Trauma is a leading cause of mortality and morbidity worldwide.
+A cornerstone in trauma quality improvement programs is multidisciplinary mortality and morbidity reviews.
+The purpose of these reviews is to identify opportunities for improvement and corresponding corrective actions.
 Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this cohort.
@@
 ### Methods
-A registry-based study of trauma patients admitted to the ICU between 2014-2023 at the Karolinska University Hospital in Solna, Sweden, who were reviewed regarding the presence of opportunities for improvement in a morbidity and mortality review.
+A registry-based study of trauma patients admitted to the ICU between 2014-2023 at the Karolinska University Hospital in Solna, Sweden, who were reviewed regarding the presence of opportunities for improvement in a mortality and morbidity review.
 Univariate and multivariable logistic regression was used to determine associations between selected patient and process factors and opportunities for improvement.
@@
 ### Results
-opportunities for improvement was identified in 143 (9,9%) out of 1449 patients.
-Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with increased odds of opportunities for improvement.
+Opportunities for improvement was identified in 143 (9,9%) out of 1449 patients.
+Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with opportunities for improvement.
 
 ### Conclusion
-Several patient and process factors were found to be associated with opportunities for improvement, indicating that patients who were relatively more stable upon arrival, patients with prolonged ICU stay, patients with prolonged mechanical ventilation, and patients who were alive 30 days after injury had higher odds of opportunities for improvement.
+Several patient and process factors were found to be associated with opportunities for improvement.
+Notably, patients who were relatively more stable upon arrival and patients with prolonged ICU stays had the highest odds of opportunities for improvement.
+This underscores the necessity of concentrating more efforts on these specific patient groups.
 
 # Introduction
+Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60][@6].
+Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and can result in complications or death [@64][@69].
+Although these complications are often expected, the complexity of ICU care necessitates providers to be highly adaptable, which makes the setting prone to mistakes and errors.
+Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42].
+The identification of these help derive corrective action plans to improve the quality of trauma care.
+Previous research on opportunities for improvement in trauma care has often focused on the initial management, identifying areas such as airway and ventilator management, fluid resuscitation, delays to treatment and judgement errors as common opportunities for improvement [@72][@73][@74].
 
-Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of morbidity and mortality worldwide [@60, @6].
-Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and result in complications or death [@64, @69].
-While these complications are often expected, the complexity of ICU-care requires providers to be highly adaptable and makes it prone to mistakes and errors.
-Peer review is a cornerstone of trauma quality improvement programs to identify preventable causes of morbidity and mortality in trauma patient [@70], referred to as opportunities for improvement [*ref* - 10 i texten].
-Previous research on opportunities for improvement in trauma care has often focused on the initial management and have identified airway and ventilator management, fluid resuscitation, delays to treatment and judgement errors as common areas for opportunities for improvement [*ref 10*].
-
-Prior research has focused on classifying errors in preventable trauma deaths, with clinical performance errors, often stemming from a lack of knowledge, being identified as the most common type, particularly in the emergency department [@62].
-However, reporting of errors in trauma resuscitation varies between settings and studies [@63].
-As a results, there is a notable gap in research regarding error reporting practices in trauma care — especially for trauma patients admitted to the ICU [@12, @64].
 Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their medical complexity, making it difficult to establish the standard of care for each individual.
 Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9].
-While numerous studies have explored various facets of trauma care enhancement, research specifically centered on opportunities for improvement in ICU-admitted trauma patients remains limited [@64].
-The aim of this study was to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
+While previous studies have explored various facets of trauma care, including the ICU-setting [@50], there is a notable gap in research regarding error reporting practices in trauma care - especially for trauma patients admitted to the ICU [@64][@12].
+Understanding and addressing this gap is essential for improving patient outcomes and optimizing trauma care protocols.
+Therefore, this study aims to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
 
@@
 
-We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which is included in the Swedish Trauma registry (SweTrau), and the hospital's local trauma care quality database.
+We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which is included in the Swedish Trauma registry, and the hospital's local trauma care quality database.
 
@@
 
+Table 1 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.
+
+```{r, include=FALSE}
+#source("functions/ofi_categories.R")
+```
+
 ## Participants
 
-The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9.
+The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score, as well as patients admitted without trauma team activation but found to have an Injury Severity Score of more than 9.
 We included all patients who had been included in the opportunities for improvement screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU.
@@
 Patients under 15 years of age were not included in the study as their clinical and review pathway differ compared to adults.
+Patients who were dead on arrival were excluded as opportunities for improvement among these patients is not relevant for the aim of this study.
 
@@
 
-When conducting logistic regression analysis, the number of events per variable (EPV) equal to 10 is a rule of thumb.
-If the EPV falls below this threshold, the results for regression analyses should be interpreted with caution [@45].
+When conducting logistic regression analysis, the number of events per variable equal to 10 is a rule of thumb.
+If the events per variable falls below this threshold, the results for regression analyses should be interpreted with caution [@45].
 
@@
 
-Keeping this guideline in mind, we selected independent variables for the analysis.
-To achieve this, we converted RTS, ISS, and time to first CT into continuous variables, while categorical variables were divided into as few levels as possible.
+We categorized and converted the independent variables according to the aforementioned guideline of events per variable to achieve an acceptable multivariate analysis.
+Specifically, Revised Trauma Score, Injury Severity Score, and time to first computed tomography were treated as continuous variables, while categorical variables were divided into as few levels as possible.
+
+### Patient and process factors
+
+The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion.
+Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality.
+Sex was categorized as male sex or female sex.
+Mechanical ventilation was stratified into “Not intubated”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”.
+On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday.
+ICU length of stay was categorized as “≤ 7 days” or “> 7 days”.
+American Society of Anesthesiologists (ASA) preinjury was divided into “ASA 1-2” and “ASA 3-6”.
+30-day mortality was determined based on whether the patient was deceased or alive 30 days after injury.
+
+The continuous variables in the analysis were age in years, Revised Trauma Score, Injury Severity Score, and time to first computed tomography in minutes.
+Revised Trauma Score is a measure of severity based on prehospital or emergency department values.
+In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Come Scale scores were missing from the emergency department records, prehospital values were used instead.
+If the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero.
+Subsequently, for the Revised Trauma Score variable, each of the variables Glasgow Coma Scale, systolic blood pressure, and respiratory rate were converted to Revised Trauma Score codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].
 
@@
 
-### Patient and process factors
-
-The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion.Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality.
-Sex was categorized as male sex or female sex.
-Mechanical ventilation was stratified into “Not intubated”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”.
-On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday.
-ICU length of stay was categorized as “≤ 7 days” or “> 7 days”.
-ASA preinjury was divided into “ASA 1-2” and “ASA 3-6”.
-30-day mortality was determined based on whether the patient was deceased or alive 30 days after injury.
-
-The continuous variables in the analysis were age in years, Revised Trauma Score (RTS), ISS, and time to first computed tomography in minutes.
-RTS is a measure of severity based on prehospital or emergency department values.
-In instances where hospital values for systolic blood pressure (SBP), respiratory rate (RR), or Glasgow Come Scale (GCS) scores were missing from the emergency department records, prehospital values were used instead.
-If the patient was intubated, it was assumed that their GCS was 3 and RR was 0.
-Subsequently, for the RTS variable, each of the variables GCS, SBP, and RR were converted to RTS codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].
-
 ## Statistical analysis
 
-We conducted complete case analysis after handling missing values for SBP, RR and GCS as described above.
+We conducted complete case analysis after handling missing values for systolic blood pressure, respiratory rate and Glasgow Coma Scale as described above.
 Pearson’s Chi-squared test and the Wilcoxon rank sum test were used to calculate the significance for categorical and continuous variables respectively.
-Continuous variables are presented using mean values and standard deviations,  and categorical variables are presented using the number of patients within each subgroup and the corresponding subgroup percentages.
+Continuous variables are presented using mean values and standard deviations, and categorical variables are presented with the number of patients in each subgroup and their respective percentages.
 
 Univariate logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and opportunities for improvement.
-Odds ratios (OR) with associated 95% confidence intervals were usecalculated.
+Odds ratios (OR) with associated 95% confidence intervals were calculated.
 A p-value of less than 0.05 considered indicative of a significant difference.
 For continuous variables, ORs were derived by evaluating each unitary step increment.
-A subgroup analysis of patients who were alive 30 days after the injury was conducted to reduce impact from death as a cause of the association between opportunities for improvement and patient and process factors.
+A subgroup analysis of patients who were alive and dead 30 days after the injury was conducted to reduce impact from death as a cause of the association between opportunities for improvement and patient and process factors.
 All statistical analyses were conducted using R version 4.4 [@19].
@@
 for (i in 1:(length(vert) - 1)) {
-  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert", arrow = small_arrow) %>%
     print
@@
 
-connectGrob(vert$eligible, excluded, type = "L")
-connectGrob(vert$included, excluded1, type = "L")
+connectGrob(vert$eligible, excluded, type = "L", arrow = small_arrow)
+connectGrob(vert$included, excluded1, type = "L", arrow = small_arrow)
 
@@
 Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study.
+The variable with the highest number of missing values was Revised Trauma Score, with 207 instances of missing data.
 
@@
 
-Table 1 present sample characteristics.
-Most patients were male 1,096 (76%) vs 353 (24%) female patients.
-There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than seven days.
-The number of patients who arrived outside working hours was 1,042 (72%).
-A total of 1,064 (73%) patients were equal to or less than seven days in the ICU, and 385 (27%) patients were more than seven days in the ICU.
+Table 2 present sample characteristics.
+Most patients were male 1,096 (76%) vs 353 (24%).
+There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than 7 days.
+The number of patients arriving outside working hours was 1,042 (72%).
+A total of 1,064 (73%) patients were equal to or less than 7 days in the ICU vs 385 (27%) patients whose ICU-stay exceeded 7 days.
 A total of 1,080 (75%) patients had an ASA score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA score of 3-6 before arrival to the hospital.
@@
 When comparing patients with opportunities for improvement to those without, differences emerged across various factors.
-Specifically, individuals with opportunities for improvement were less likely to require intubation, had higher RTS scores, experienced shorter time to initial CT scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
+Specifically, individuals with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
 
@@
 
-Table 2 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement.
+Table 3 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement.
 In the unadjusted analysis, the factors that were significantly associated with the presence of opportunities for improvement were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.
 
-In the adjusted analysis the following factors demonstrated significant association with increasing odds of opportunities for improvement: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
+In the adjusted analysis the following factors demonstrated significant association with opportunities for improvement: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
 
-Sex, age, ISS, time to first CT, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
+Sex, age, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
 
@@
 ```{r, echo=FALSE}
-table3alive_merge
+table3b_merge
 ```
 
-Table 3 shows the unadjusted and adjusted associations of opportunities for improvement and the selected patient and process factors among the patients who were alive 30 days after injury.
-In the unadjusted analysis > 7 days in the ICU was significantly associated with opportunities for improvement.
-In the adjusted analysis increased RTS-score was significantly associated with increased odds of opportunities for improvement.
+Table 4 shows the adjusted associations of opportunities for improvement and the selected patient and process factors among the patients who were alive and dead respectively 30 days after injury.
+Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
+More than 7 days in the ICU was significantly associated with increased odds of opportunities for improvement among patients who were deceased 30 days after injury `r inline_text(table3bdead, variable = daysinICU, level = "> 7 days")`.
 
@@
 
-The key results from this study are that higher RTS, > 7 days length of stay in the ICU and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
-Age, sex, mechanical ventilation, ISS, time to first CT, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in either the adjusted or the unadjusted analyses.
-In the subgroup analysis of patients who were alive 30 days after injury, no patient or process factor was significantly associated with opportunities for improvement in the adjusted analysis.
+The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
+More than 7 days on mechanical ventilation was associated with decreased odds of opportunities for improvement.
+Age, sex, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the adjusted nor the unadjusted analyses.
+In the analysis of patients who were alive and deceased 30 days after injury increased Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
+Among the patients who deceased 30 days after injury, more than 7 days in the ICU was significantly associated with higher odds of opportunities for improvement.
 
-We found that higher RTS was associated with increased odds of opportunities for improvement.
-Previous research has shown that patients with relatively better vital parameter values, had a significant association with increased odds of opportunities for improvement in the unadjusted analysis, but not in the adjusted analysis [@12].
-These findings could be because patients with lower RTS are often in critical condition upon arrival — upon which emergency interventions are performed immediately.
-Conversely, patients with higher RTS, indicating relative stability in this patient group upon arrival, may experience delayed interventions.
+Our study found a significant association between higher Revised Trauma Score and higher odds of opportunities for improvement, indicating that among very severely injured patients that are admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement.
+Patients with lower Revised Trauma Score are often in critical condition upon arrival — upon which emergency interventions are performed immediately.
+In contrast, patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring.
+This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have increased odds of opportunities for improvement [@12].
 
-Similarly, prolonged ICU stay was found to be significantly associated with opportunities for improvement.
-There are two things to consider regarding this variable.
-Firstly, patients requiring extended ICU care are often those with complex medical conditions, severe injuries, or critical illness.
-As a result, their prolonged stay in the ICU may signify a higher baseline risk for experiencing complications, increasing odds of opportunities for improvement.
-Additionally, prolonged ICU stay increases likelihood of medical errors and adverse events to occur [*REF*].
-The ICU environment, while designed to provide advanced medical care, can also be fraught with complexities, including high patient acuity, numerous interventions, and a fast-paced workflow.
-The longer a patient remains in the ICU, the greater the cumulative exposure to potential errors.
+Prolonged ICU stay was also found to be significantly associated with opportunities for improvement.
+This variable warrants careful consideration for two primary reasons.
+Firstly, patients requiring extended ICU care are often those with complex medical conditions or severe injuries.
+As a result, their prolonged ICU stay may signify a higher baseline risk for experiencing complications, increasing odds of opportunities for improvement.
+The ICU environment, while designed to provide advanced medical care, can also be fraught with complexities, including high patient acuity, numerous interventions, and a fast-paced workflow, all contributing to a higher cumulative exposure to potential errors the longer a patient remains in the ICU.
+Additionally, complications or opportunities for improvement might themselves lead to longer ICU stays.
 Conversely, shorter ICU stays may indicate patient demise before potential mistakes could occur, particularly in cases where mortality was anticipated rather than preventable.
-This finding is similar to previous research demonstrating prolonged ICU stays as a risk factor for increased morbidity and mortality in trauma patients [@22, @26].
+This finding is similar to previous research demonstrating prolonged ICU stays as a risk factor for increased mortality and morbidity in trauma patients [@22, @26].
 Furthermore, it also corresponds with a retrospective cohort indicating that lower initial illness severity and younger age are associated with earlier discharge, further supporting the link between higher odds of opportunities for improvement and prolonged ICU stays [@25].
@@
 The cause of the organ dysfunction comes from increased exposure to iatrogenic factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions.
-[@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and ISS in our study.
-These factors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and ISS.
+[@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and Injury Severity Score in our study.
+These factors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and Injury Severity Score.
 
@@
 
-We also found that mechanical ventilation more than seven days was significantly associated with decreased odds of opportunities for improvement.
+We found that mechanical ventilation more than 7 days was significantly associated with decreased odds of opportunities for improvement.
 Previous research has shown that prolonged mechanical ventilation is associated with increased mortality.
-This is because mechanical ventilation is associated with severe complications and failure of weaning further increases length of mechanical ventilation as well as ICU length of stay [@65, @66].
-Our finding of this association may thus be because of the increased rate of complications and mortality associated with prolonged mechanical ventilation.
-*skriv om så det gäller minskad OFI*
+This is because mechanical ventilation is associated with severe complications and failure of weaning further increases length of mechanical ventilation as well as ICU length of stay [@65][@66].
+Our finding of decreased odds of OFI may be attributed to the fact that, despite the increased rate of complications and mortality associated with prolonged mechanical ventilation, it was the appropriate intervention for these patients.
 
-Several limitations need to be acknowledged in our study.
+One of the strengths of this study is its reliance on registry-based data, ensuring a high level of data quality and reliability.
+The use of pre-defined variables standardized their definitions, enhancing the potential for similar studies to replicate these findings.
+Additionally, the study’s specific inclusion criteria facilitated the formation of a well-defined cohort, further contributing to the reliability and reproducibility of the study.
+
+However, several limitations need to be acknowledged in our study.
 This was a single-center study and primarily provide insights into the context of the Karolinska University Hospital in Solna.
-Another limitation is the selection of patients for the morbidity and mortality review, which heavily relied on the local audit filters.
-The use of audit filters for patient selection for the morbidity and mortality review is associated with high false positive rates, ranging from 24% to 80% [@38].
+Another limitation is the selection of patients for the mortality and mortbidity review, which heavily relied on the local audit filters.
+The use of audit filters for patient selection for the mortality and morbidity review is associated with high false positive rates, ranging from 24% to 80% [@38].
 Even though the use of audit filters is associated with high false positive rates, there is still a risk of misplacement where some patients with opportunities for improvement may have been overlooked.
-Furthermore, the handling of the variables to the RTS score – namely RR, SBP, and GCS – introduced potential sources of variability.
+Furthermore, the handling of the variables to the Revised Trauma Score – namely respiratory rate, systolic blood pressure, and Glasgow Coma Scale – introduced potential sources of variability.
 The use of prehospital, emergency department, or intubation-based values for these variables may have affected data accuracy and reproducibility.
@@
 # Conclusion
-The findings of this study revealed a significant association between higher RTS scores, ICU stay exceeding seven days, mechanical ventilation exceeding seven days, and survival 30 days after injury with increased odds of opportunities for improvement.
-These factors represent viable targets for new initiatives to further improve the quality of trauma care.
+
+The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay exceeding 7 days, and survival 30 days after injury, as well as a significant association between decreased odds of opportunities for improvement and more than 7 days on mechanical ventilation.
+The results highlight the importance of prioritizing these groups of patients for closer monitoring and implementing targeted interventions to prevent errors in care.
+Further research on opportunities for improvement among ICU-admitted patients is needed to expand our understanding of this aspect of trauma care.
+The primary limitation of this study was its restricted sample size, highlighting the need for larger-scale investigations.
+Conducting multi-center studies encompassing extensive databases or merging multiple datasets could provide more comprehensive insights into opportunities for improvement and contribute to enhancing Trauma Quality Improvement Programs on a broader scale.
+However, it is important to acknowledge that multi-center studies may compromise internal validity, reducing their applicability to specific centers.
+
+Building upon the findings of this study, future research should focus on specific variables such as ICU length of stay and the individual assessment of vital parameters like Glasgow Coma Scale, respiratory rate, and systolic blood pressure, rather than relying solely on composite scores like Revised Trauma Score.
+Analyzing these vital signs separately may reveal which parameter holds greater predictive value for opportunities for improvement.
+Also, qualitative studies could provide an in-depth understanding of why physiologically more stable patients are more likely to have opportunities for improvement.
+
+Given the significant association between higher Revised Trauma Score and increased odds of opportunities for improvement, attention should be directed towards trauma patients who appear stable upon admission but may deteriorate over time.
+Future research should aim to identify early indicators of worsening conditions in these patients, facilitating more targeted resource allocation and intervention strategies.
+
+In contrast, Injury Severity Score did not demonstrate a correlation with opportunities for improvement in this study.
+Subsequent studies could explore alternative scoring systems such as New Injury Severity Score, Revised Trauma Score, and ASA scores, or combinations of physiological and anatomical scoring systems to determine the most effective predictor of opportunities for improvement in trauma patients.
+
+Overall, future research endeavors should strive to address these gaps in knowledge to optimize trauma care outcomes and contribute to the ongoing improvement of trauma quality improvement initiatives.
 

### 2024-06-05 functions/flowchart.R

@@ -72,3 +72,4 @@ connectGrob(vert$included, excluded1, type = "L", arrow = small_arrow)
 vert
 excluded
 excluded1
+

### 2024-06-05 functions/ofi_categories.R

@@ -1,0 +1,326 @@
+suppressPackageStartupMessages({
+  library(rofi)
+  library(dplyr)
+  library(readr)
+  library(stringr)
+  library(hms)
+  library(lubridate)
+})
+
+################################################################################
+#' Prepare original dataframe
+prepare_data <- function(df) {
+  # drop irrelevant
+  df <- df %>% 
+    select(pt_age_yrs, pt_Gender, pt_asa_preinjury, inj_intention, inj_mechanism,
+           inj_dominant, NISS, starts_with("AISCode_"), starts_with("ICD_"),
+           DateTime_ArrivalAtHospital, Problemomrade_.FMP, ofi_review, arrival)
+  
+  # create x_sex factor of sex - `Male`/`Female`
+  df$x_sex <- factor(df$pt_Gender, levels = c(1, 2), labels = c("Male", "Female"))
+  attr(df$x_sex, "label") <- "Sex"
+  
+  #clean age into x_age
+  df$x_age <- ifelse(df$pt_age_yrs > 150, NA, df$pt_age_yrs)
+  attr(df$x_age, "label") <- "Age"
+  
+  # group pt_asa_preinjury into x_asa
+  df$x_asa <- factor(
+    case_when(
+      df$pt_asa_preinjury %in% c(1, 2) ~ "a",
+      df$pt_asa_preinjury == 3 ~ "b",
+      df$pt_asa_preinjury %in% c(4, 5, 6) ~ "c",
+      TRUE ~ NA
+    ),
+    levels = c("a", "b", "c"),
+    labels = c("Minor", "Moderate", "Severe")
+  )
+  attr(df$x_asa, "label") <- "Preinjury comorbidities"
+  
+  # attempt to calculate exposure group PRN
+  source("functions/icd.R")
+  df$x_group <- as.factor(ifelse(is.na(df$inj_intention) | df$inj_intention > 3,
+                                 calc_intent(na.omit(unname(select(df, starts_with("ICD_"))))) %||% df$inj_intention,
+                                 df$inj_intention))
+  attr(df$x_group, "label") <- "Injury intent"
+  EXPOSED <- "Intentional"
+  UNEXPOSED <- "Unintentional"
+  
+  # df$x_group <- as.factor(ifelse(is.na(df$inj_mechanism) | df$inj_mechanism > 12,
+  #                       calc_moi(na.omit(unname(select(df, starts_with("ICD_"))))) %||% df$inj_mechanism,
[…]

### 2024-06-05 functions/table1.R

@@ -62,14 +62,14 @@ ofi$Age <- ofi$pt_age_yrs
 
 #Intubation 
 ofi$Intubation1 <- ifelse(ofi$pre_intubated == 1, "Intubation",
-                          ifelse(ofi$pre_intubated == 2, "No intubation",  
+                          ifelse(ofi$pre_intubated == 2, "Not intubated",  
                                  ifelse(ofi$pre_intubated == 999, "Unknown",
                                         ifelse(ofi$ed_intubated == 1, "Intubation",
-                                               ifelse(ofi$ed_intubated == 2, "No intubation",  
+                                               ifelse(ofi$ed_intubated == 2, "Not intubated",  
                                                       ifelse(ofi$ed_intubated == 999, "Unknown", "Unknown"))))))
 
 #Intubation combined with ventilator days 
-ofi$Intubation <- ifelse(ofi$Intubation1 == "No intubation", "No intubation",
+ofi$Intubation <- ifelse(ofi$Intubation1 == "Not intubated", "Not intubated",
                          ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days ==  0, "Mechnical ventilation 1-7 days",
                                 ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days %in% 1:7, "Mechanical ventilation 1-7 days",
                                        ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days > 7, "Mechanical ventilation > 7 days", 
@@ -185,7 +185,7 @@ table1$Intubation <- ifelse(is.na(table1$Intubation), "Unknown", table1$Intubati
 table1 <- na.omit(table1)
 
 table2 <- table1 %>%
-  mutate(Intubation = factor(Intubation, levels = c("No intubation", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown"))) %>%
+  mutate(Intubation = factor(Intubation, levels = c("Not intubated", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown"))) %>%
   tbl_summary(by = OpportunityForImprovement,
               type = list(OnDuty ~ "dichotomous"),
               label = list(RTS = "Revised Trauma Score",
@@ -212,6 +212,6 @@ table2 <- table1 %>%
   add_overall(last = TRUE) %>% 
   #add_p() %>%
   #bold_p(t=0.05) %>%
[…]

### 2024-06-05 functions/table2.R

@@ -6,7 +6,7 @@ tablereg <- ofi %>%
 tablereg <- na.omit(tablereg)
 
 tablereg$Intubation <- ifelse(is.na(tablereg$Intubation), "Unknown", table1$Intubation)
-tablereg$Intubation <- fct_relevel(tablereg$Intubation, "No intubation", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown")
+tablereg$Intubation <- fct_relevel(tablereg$Intubation, "Not intubated", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown")
 tablereg$daysinICU <- fct_relevel(tablereg$daysinICU, "≤ 7 days", "> 7 days")
 tablereg$Survival <- fct_relevel(tablereg$Survival, "Dead", "Alive")
 
@@ -26,8 +26,7 @@ table3a <- tbl_uvregression(data = tablereg,
                               TimeFCT = "Time to first CT"
                             )) %>%
   bold_labels() %>%
-  bold_p(t = 0.05) %>%
-  hide_n = TRUE
+  bold_p(t = 0.05)
 
 #print(table3a)
 
@@ -44,15 +43,15 @@ table3b <- tbl_regression(adjusted_table,
                                        ASApreinjury = "ASA preinjury",
                                        TimeFCT = "Time to first CT")) %>%
   bold_labels() %>%
-  bold_p(t = 0.05) %>%
-  hide_n = TRUE
+  bold_p(t = 0.05)
 
 # print(table3b)
 
 # Merging Tables
 table3_merge <- tbl_merge(tbls = list(table3a, table3b),
                           tab_spanner = c("**Unadjusted**", "**Adjusted**")) %>%
-  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 2. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement (N = 1200).</div>")
+ # modify_column_merge(pattern = "{OR} ({ci}) {p.value}") %>%
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 3. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement (N = 1200).</div>")
 
 
 print(table3_merge)

### 2024-06-05 functions/table3.R

@@ -31,7 +31,7 @@ tablereg1$daysinICU <- fct_relevel(tablereg1$daysinICU, "≤ 7 days", "> 7 days"
 #Creating linear regression 
 adjusted_table1 <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury, family = binomial, data = tablereg1) 
 
-table3alive <- tbl_regression(adjusted_table1,
+table3aalive <- tbl_regression(adjusted_table1,
                           exponentiate = TRUE, 
                           label = list(RTS = "Revised Trauma Score",
                                        daysinICU = "Days in the ICU",
@@ -40,8 +40,7 @@ table3alive <- tbl_regression(adjusted_table1,
                                        ASApreinjury = "ASA preinjury",
                                        TimeFCT = "Time to first CT")) %>%
   bold_labels() %>%
-  bold_p(t = 0.05) %>%
-  hide_n = TRUE
+  bold_p(t = 0.05)
 
 #Creating with patients who died
 ofi_dead <- subset(tablereg, subset = (Survival == "Dead"))
@@ -54,7 +53,7 @@ tablereg2$daysinICU <- fct_relevel(tablereg2$daysinICU, "≤ 7 days", "> 7 days"
 
 adjusted_table2 <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury, family = binomial, data = tablereg2) 
 
-table3dead <- tbl_regression(adjusted_table2,
+table3bdead <- tbl_regression(adjusted_table2,
                                exponentiate = TRUE, 
                                label = list(RTS = "Revised Trauma Score",
                                             daysinICU = "Days in the ICU",
@@ -63,14 +62,14 @@ table3dead <- tbl_regression(adjusted_table2,
                                             ASApreinjury = "ASA preinjury",
                                             TimeFCT = "Time to first CT")) %>%
   bold_labels() %>%
-  bold_p(t = 0.05) %>%
-  hide_n = TRUE
+  bold_p(t = 0.05)
 
 # Merging Tables
-table3_merge <- tbl_merge(tbls = list(table3alive, table3dead),
+table3b_merge <- tbl_merge(tbls = list(table3aalive, table3bdead),
                           tab_spanner = c("**Alive**", "**Dead**")) %>%
[…]

### 2024-06-09 manuscript2.Rmd
9/6

@@
 ### Results
-Opportunities for improvement was identified in 143 (9,9%) out of 1449 patients.
-Mechanical ventilation > 7 days (OR 0.35; 95% CI 0.10, 0.89; p=0.05), Revised Trauma Score (OR 1.21; 95% CI 1.08, 1.37; p=0,002), > 7 days in the ICU (OR 1.93; 95% CI 1.23, 3.03; p=0.004) and survival 30 days after injury (OR 3.29; 95% CI 1.60, 7.54; p=0.002) were significantly associated with opportunities for improvement.
+Opportunities for improvement was identified in 143 (9,9%) out of 1449 patients.Revised Trauma Score , > 7 days in the ICU  and survival 30 days after injury  were significantly associated with increased opportunities for improvement.
 
@@
 # Introduction
-Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60][@6].
-Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and can result in complications or death [@64][@69].
+Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60; @6].
+Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and can result in complications or death [@64, @69].
 Although these complications are often expected, the complexity of ICU care necessitates providers to be highly adaptable, which makes the setting prone to mistakes and errors.
@@
 The identification of these help derive corrective action plans to improve the quality of trauma care.
-Previous research on opportunities for improvement in trauma care has often focused on the initial management, identifying areas such as airway and ventilator management, fluid resuscitation, delays to treatment and judgement errors as common opportunities for improvement [@72][@73][@74].
+Previous research on opportunities for improvement in trauma care have been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
 
@@
 Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9].
-While previous studies have explored various facets of trauma care, including the ICU-setting [@50], there is a notable gap in research regarding error reporting practices in trauma care - especially for trauma patients admitted to the ICU [@64][@12].
+While previous studies have explored various facets of trauma care, including the ICU-setting [@50], there is a notable gap in research regarding error reporting practices in trauma care - especially for trauma patients admitted to the ICU [@64; @12].
 Understanding and addressing this gap is essential for improving patient outcomes and optimizing trauma care protocols.
@@
 The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database.
-Before 2017 a non-systematic selection from clinicians involved in the opportunities for improvement process was done by filtering for patients that were thought to have a high risk for opportunities for improvement, but starting in 2017 the care of all patients is screened.
+Prior to 2017 a semi-systematic selection off clinicians involved in the opportunities for improvement process was done by filtering for patients who were thought to have a high risk for opportunities for improvement.
+However, starting in 2017, the care of all patients is screened using audit filters.
 
@@
 ```{r, include=FALSE}
-#source("functions/ofi_categories.R")
+source("functions/ofi_categories.R")
+```
+```{r, echo=FALSE}
+table2
 ```
@@
 
-In the adjusted analysis the following factors demonstrated significant association with opportunities for improvement: mechanical ventilation > 7 days `r inline_text(table3b, variable = Intubation, level = "Mechanical ventilation > 7 days")`, Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
+In the adjusted analysis the following factors demonstrated significant association with opportunities for improvement: Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
 
-Sex, age, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
+Sex, age, mechanical ventilation, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
 
@@
 Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
-More than 7 days in the ICU was significantly associated with increased odds of opportunities for improvement among patients who were deceased 30 days after injury `r inline_text(table3bdead, variable = daysinICU, level = "> 7 days")`.
 
@@
 The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
-More than 7 days on mechanical ventilation was associated with decreased odds of opportunities for improvement.
-Age, sex, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the adjusted nor the unadjusted analyses.
+Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the adjusted nor the unadjusted analyses.
 In the analysis of patients who were alive and deceased 30 days after injury increased Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
@@
 
-We found that mechanical ventilation more than 7 days was significantly associated with decreased odds of opportunities for improvement.
-Previous research has shown that prolonged mechanical ventilation is associated with increased mortality.
-This is because mechanical ventilation is associated with severe complications and failure of weaning further increases length of mechanical ventilation as well as ICU length of stay [@65][@66].
-Our finding of decreased odds of OFI may be attributed to the fact that, despite the increased rate of complications and mortality associated with prolonged mechanical ventilation, it was the appropriate intervention for these patients.
-
 One of the strengths of this study is its reliance on registry-based data, ensuring a high level of data quality and reliability.
@@
 
-The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay exceeding 7 days, and survival 30 days after injury, as well as a significant association between decreased odds of opportunities for improvement and more than 7 days on mechanical ventilation.
+The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay exceeding 7 days, and survival 30 days after injury.
 The results highlight the importance of prioritizing these groups of patients for closer monitoring and implementing targeted interventions to prevent errors in care.

### 2024-06-09 functions/ofi_categories.R

@@ -1,326 +1,371 @@
-suppressPackageStartupMessages({
-  library(rofi)
-  library(dplyr)
-  library(readr)
-  library(stringr)
-  library(hms)
-  library(lubridate)
-})
+#INSTALLING PACKAGES
+#devtools::install_github("martingerdin/noacsr")
+#devtools::install_github("martingerdin/rofi")
+library(dotenv)
+library(noacsr)
+library(rofi)
+#noacsr::source_all_functions()
+data <- import_data()
 
-################################################################################
-#' Prepare original dataframe
-prepare_data <- function(df) {
-  # drop irrelevant
-  df <- df %>% 
-    select(pt_age_yrs, pt_Gender, pt_asa_preinjury, inj_intention, inj_mechanism,
-           inj_dominant, NISS, starts_with("AISCode_"), starts_with("ICD_"),
-           DateTime_ArrivalAtHospital, Problemomrade_.FMP, ofi_review, arrival)
-  
-  # create x_sex factor of sex - `Male`/`Female`
-  df$x_sex <- factor(df$pt_Gender, levels = c(1, 2), labels = c("Male", "Female"))
-  attr(df$x_sex, "label") <- "Sex"
-  
-  #clean age into x_age
-  df$x_age <- ifelse(df$pt_age_yrs > 150, NA, df$pt_age_yrs)
-  attr(df$x_age, "label") <- "Age"
-  
-  # group pt_asa_preinjury into x_asa
-  df$x_asa <- factor(
-    case_when(
-      df$pt_asa_preinjury %in% c(1, 2) ~ "a",
-      df$pt_asa_preinjury == 3 ~ "b",
-      df$pt_asa_preinjury %in% c(4, 5, 6) ~ "c",
-      TRUE ~ NA
-    ),
-    levels = c("a", "b", "c"),
-    labels = c("Minor", "Moderate", "Severe")
-  )
-  attr(df$x_asa, "label") <- "Preinjury comorbidities"
-  
-  # attempt to calculate exposure group PRN
-  source("functions/icd.R")
-  df$x_group <- as.factor(ifelse(is.na(df$inj_intention) | df$inj_intention > 3,
-                                 calc_intent(na.omit(unname(select(df, starts_with("ICD_"))))) %||% df$inj_intention,
-                                 df$inj_intention))
-  attr(df$x_group, "label") <- "Injury intent"
-  EXPOSED <- "Intentional"
-  UNEXPOSED <- "Unintentional"
-  
[…]

### 2024-06-09 functions/table1.R

@@ -1,173 +1,3 @@
-#INSTALLING PACKAGES
-#devtools::install_github("martingerdin/noacsr")
-#devtools::install_github("martingerdin/rofi")
-library(dotenv)
-library(noacsr)
-library(rofi)
-#noacsr::source_all_functions()
-data <- import_data()
-
-merged.data <- merge_data(data)
-merged.data$ofi <- create_ofi(merged.data)
-
-
-#install.packages("dplyr")
-library(dplyr)
-#install.packages("gtsummary")
-library(gtsummary)
-library(tidyverse)
-library(officer)
-
-
-#FLOWCHART: included/excluded
-#install.packages("Gmisc")
-library(Gmisc, quietly = TRUE)
-library(glue)
-#install.packages("htmlTable")
-library(htmlTable)
-library(grid)
-library(magrittr)
-
-##CLEANING DATA
-subdat <- merged.data %>%
-  select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
-         res_survival, pre_intubated, ed_intubated, dt_ed_first_ct, ISS, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
-         host_care_level, hosp_vent_days, pt_asa_preinjury, pre_gcs_sum, 
-         pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat, iva_dagar_n)
-
-#Converting subdat$ofi to logical so subset can be used 
-subdat$ofi <- ifelse(subdat$ofi == "Yes", TRUE, FALSE)
-
-#Only those in IVA
-iva <- subset(subdat, subset = (host_care_level == 5))
-
-#Removing pt_yrs < 15
-adult <- subset(iva, subset = (pt_age_yrs > 14))
-
-#Deceased on arrival 
-alive <- subset(adult, subset = (Fr1.12 == 2 | is.na(Fr1.12)))
-
-#Removing ofi = NA            
-ofi <- alive %>% subset(!is.na(ofi))
-
-
-#DEFINING VARIABLES FOR TABLE 1 
-#Gender
-ofi$Sex <- ifelse(ofi$pt_Gender == 1, "Male", 
-                  ifelse(ofi$pt_Gender == 2, "Female", 
[…]

### 2024-06-09 functions/table2.R

@@ -3,12 +3,13 @@
 tablereg <- ofi %>% 
   select(Sex, Age, Intubation, RTS, ISS,TimeFCT, OnDuty, daysinICU, TimeFCT, 
          ASApreinjury, Survival, OpportunityForImprovement1)
-tablereg <- na.omit(tablereg)
 
 tablereg$Intubation <- ifelse(is.na(tablereg$Intubation), "Unknown", table1$Intubation)
 tablereg$Intubation <- fct_relevel(tablereg$Intubation, "Not intubated", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown")
 tablereg$daysinICU <- fct_relevel(tablereg$daysinICU, "≤ 7 days", "> 7 days")
 tablereg$Survival <- fct_relevel(tablereg$Survival, "Dead", "Alive")
+
+tablereg <- na.omit(tablereg)
 
 # Unadjusted Table
 table3a <- tbl_uvregression(data = tablereg,
@@ -51,7 +52,7 @@ table3b <- tbl_regression(adjusted_table,
 table3_merge <- tbl_merge(tbls = list(table3a, table3b),
                           tab_spanner = c("**Unadjusted**", "**Adjusted**")) %>%
  # modify_column_merge(pattern = "{OR} ({ci}) {p.value}") %>%
-  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 3. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement (N = 1200).</div>")
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 3. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement (N = 1449).</div>")
 
 
 print(table3_merge)

### 2024-06-09 functions/table3.R

@@ -40,6 +40,7 @@ table3aalive <- tbl_regression(adjusted_table1,
                                        ASApreinjury = "ASA preinjury",
                                        TimeFCT = "Time to first CT")) %>%
   bold_labels() %>%
+  add_n() %>%
   bold_p(t = 0.05)
 
 #Creating with patients who died
@@ -68,7 +69,7 @@ table3bdead <- tbl_regression(adjusted_table2,
 table3b_merge <- tbl_merge(tbls = list(table3aalive, table3bdead),
                           tab_spanner = c("**Alive**", "**Dead**")) %>%
  # modify_table_styling(table3b_merge, hide_n = TRUE) %>%
-  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 4. Adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement in patients alive and dead 30 days after hospitalization (N = 953).</div>")
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 4. Adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement in patients alive and dead 30 days after hospitalization (N = 1163).</div>")
 
 # Print the merged table
 print(table3b_merge) 

### 2024-06-26 manuscript2.Rmd
26/6

@@
 ---
-title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
+title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based cohort study"
 output: "html_document"
@@
 
-## Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study
+## Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based cohort study
+## Abstract
 ### Background
@@
 Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60; @6].
-Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and can result in complications or death [@64, @69].
-Although these complications are often expected, the complexity of ICU care necessitates providers to be highly adaptable, which makes the setting prone to mistakes and errors.
+Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and result in complications or death [@64; @69].
+Although these complications are often expected, the complexity of ICU care makes the setting prone to mistakes and errors.
+
 Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42].
-The identification of these help derive corrective action plans to improve the quality of trauma care.
-Previous research on opportunities for improvement in trauma care have been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
+Previous research on opportunities for improvement in trauma care has focused on the initial resuscitation phase and show that
+*detta ska redigeras*
+been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
 
-Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their medical complexity, making it difficult to establish the standard of care for each individual.
-Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9].
-While previous studies have explored various facets of trauma care, including the ICU-setting [@50], there is a notable gap in research regarding error reporting practices in trauma care - especially for trauma patients admitted to the ICU [@64; @12].
-Understanding and addressing this gap is essential for improving patient outcomes and optimizing trauma care protocols.
-Therefore, this study aims to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
+In contrast, little is known about opportunities for improvement in the care of trauma patients in the ICU [@64; @12].
+Assessing opportunities for improvement in these patients is challenging because it is difficult to establish the standard of care for each individual and despite optimal treatment, these patients are at risk for complications and death [@9].
+This study aims to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.
 
@@
 
-## Study design
+## Study design and data sources
 
-We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which is included in the Swedish Trauma registry, and the hospital's local trauma care quality database.
+We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital's local trauma care quality database.
+The trauma registry is part of the Swedish national trauma registry SweTrau [*källa*].
 
@@
 The care of adult trauma patients is screened for opportunities for improvement using a combination of audit filters and individual review by specialized nurses.
-Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review held every 6-8 weeks.
+Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review meeting held every 6-8 weeks.
 The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database.
-Prior to 2017 a semi-systematic selection off clinicians involved in the opportunities for improvement process was done by filtering for patients who were thought to have a high risk for opportunities for improvement.
-However, starting in 2017, the care of all patients is screened using audit filters.
-
-Table 1 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.
-
-```{r, include=FALSE}
-source("functions/ofi_categories.R")
-```
-```{r, echo=FALSE}
-table2
-```
 
@@
 
-When conducting logistic regression analysis, the number of events per variable equal to 10 is a rule of thumb.
-If the events per variable falls below this threshold, the results for regression analyses should be interpreted with caution [@45].
+We included all eligible patients from the trauma registry.
 
@@
 
-We categorized and converted the independent variables according to the aforementioned guideline of events per variable to achieve an acceptable multivariate analysis.
-Specifically, Revised Trauma Score, Injury Severity Score, and time to first computed tomography were treated as continuous variables, while categorical variables were divided into as few levels as possible.
+### Outcome
+
+The outcome was the presence of any opportunity for improvement.
+An opportunity for improvement presents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice.
 
@@
 The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion.
-Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality.
+The categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, American Society of Anesthesiologists preinjury score, and 30-day mortality.
 Sex was categorized as male sex or female sex.
@@
 ICU length of stay was categorized as “≤ 7 days” or “> 7 days”.
-American Society of Anesthesiologists (ASA) preinjury was divided into “ASA 1-2” and “ASA 3-6”.
-30-day mortality was determined based on whether the patient was deceased or alive 30 days after injury.
+American Society of Anesthesiologists (ASA) preinjury score was divided into “ASA 1-2” and “ASA 3-6”.
+We categorised variables to limit the number of independent variables in the multivariable analysis to adhere to the rule of thumb of ten patients with the outcome per variable [@45].
 
 The continuous variables in the analysis were age in years, Revised Trauma Score, Injury Severity Score, and time to first computed tomography in minutes.
-Revised Trauma Score is a measure of severity based on prehospital or emergency department values.
+The Revised Trauma Score is a measure of severity based on prehospital or emergency department values.
 In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Come Scale scores were missing from the emergency department records, prehospital values were used instead.
@@
 
-### Study outcome
-
-The study’s outcome was the presence of any opportunity for improvement.
-An opportunity for improvement presents preventable errors in care with direct adverse effects on patient outcomes or recurrent deviations from safe clinical practice.
-Examples include technical error, clinical judgement error or any preventable death.
-
 ## Statistical analysis
 
-We conducted complete case analysis after handling missing values for systolic blood pressure, respiratory rate and Glasgow Coma Scale as described above.
-Pearson’s Chi-squared test and the Wilcoxon rank sum test were used to calculate the significance for categorical and continuous variables respectively.
+We conducted a complete case analysis after handling missing values for systolic blood pressure, respiratory rate and Glasgow Coma Scale as described above.
 Continuous variables are presented using mean values and standard deviations, and categorical variables are presented with the number of patients in each subgroup and their respective percentages.
 
-Univariate logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and opportunities for improvement.
+We used the Pearson’s Chi-squared test and the Wilcoxon rank sum test to compare categorical and continuous variables respectively between patients with and without opportunities for improvement.
+
+We used logistic regression to determine unadjusted and adjusted associations between patient and process factors and opportunities for improvement.
 Odds ratios (OR) with associated 95% confidence intervals were calculated.
 A p-value of less than 0.05 considered indicative of a significant difference.
-For continuous variables, ORs were derived by evaluating each unitary step increment.
-A subgroup analysis of patients who were alive and dead 30 days after the injury was conducted to reduce impact from death as a cause of the association between opportunities for improvement and patient and process factors.
+As a sensitivity analysis, we considered 30-day mortality as a potential effect modifier and therefore conducted a stratified analysis in which the associations between patient and process factors and opportunities for improvement were estimated separately in patients who survived and patients who died within 30 days.
 All statistical analyses were conducted using R version 4.4 [@19].
@@
 
-The patient demographics of this study is presented in Figure 1.
-A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded.
+A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded, as shown in Figure 1.
 Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement This left us with a total of 1,742 patients eligible for the study.
@@
 
-Table 2 present sample characteristics.
+Table 1 present sample characteristics.
 Most patients were male 1,096 (76%) vs 353 (24%).
@@
 Amongst our patients, 30-day mortality was at 286 (20%).
+When comparing patients with opportunities for improvement to those without, patients with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
 
-When comparing patients with opportunities for improvement to those without, differences emerged across various factors.
-Specifically, individuals with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
+```{r, include=FALSE}
+source("functions/ofi_categories.R")
+```
+```{r, echo=FALSE}
+table2
+```
+
+Table 2 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.
 
@@
 Table 3 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement.
-In the unadjusted analysis, the factors that were significantly associated with the presence of opportunities for improvement were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.
-
-In the adjusted analysis the following factors demonstrated significant association with opportunities for improvement: Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
-
+In the unadjusted analysis, Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`, were all significantly associated with opportunities for improvement.
+These variables remained significantly associated with opportunities for improvement in the adjusted analysis, with similar effect magnitudes: Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.
 Sex, age, mechanical ventilation, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.
@@
 
-Table 4 shows the adjusted associations of opportunities for improvement and the selected patient and process factors among the patients who were alive and dead respectively 30 days after injury.
-Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
+Table 4 shows the results of the stratified analysis with adjusted associations between patient and process factors and opportunities for improvement among the patients who were alive and dead 30 days after injury respectively.
+The Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
 
@@
 The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
-Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the adjusted nor the unadjusted analyses.
-In the analysis of patients who were alive and deceased 30 days after injury increased Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
+Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the unadjusted nor the adjusted analyses.
+In the stratified analysis of patients who were alive and dead 30 days after injury, only the Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
 Among the patients who deceased 30 days after injury, more than 7 days in the ICU was significantly associated with higher odds of opportunities for improvement.
@@
 Our study found a significant association between higher Revised Trauma Score and higher odds of opportunities for improvement, indicating that among very severely injured patients that are admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement.
-Patients with lower Revised Trauma Score are often in critical condition upon arrival — upon which emergency interventions are performed immediately.
+Patients with lower Revised Trauma Score are in a critical condition upon arrival — upon which emergency interventions are performed immediately.
 In contrast, patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring.
-This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have increased odds of opportunities for improvement [@12].
+This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
 

### 2024-06-30 manuscript2.Rmd
30/6

@@
 Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42].
-Previous research on opportunities for improvement in trauma care has focused on the initial resuscitation phase and show that
-*detta ska redigeras*
-been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
+Previous research on opportunities for improvement in trauma care have been compiled into a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
 
@@
 We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital's local trauma care quality database.
-The trauma registry is part of the Swedish national trauma registry SweTrau [*källa*].
+The trauma registry is part of the Swedish national trauma registry SweTrau [@76].
 
@@
 ```{r, echo=FALSE}
-table2
+table_ofic
 ```
@@
 The Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
+Time to first computed tomography was significantly associated with opportunities for improvement in patients who were deceased 30 days after injury `r inline_text(table3bdead, variable = TimeFCT)`.
 
@@
 
-The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay exceeding 7 days, and survival 30 days after injury.
-The results highlight the importance of prioritizing these groups of patients for closer monitoring and implementing targeted interventions to prevent errors in care.
-Further research on opportunities for improvement among ICU-admitted patients is needed to expand our understanding of this aspect of trauma care.
-The primary limitation of this study was its restricted sample size, highlighting the need for larger-scale investigations.
-Conducting multi-center studies encompassing extensive databases or merging multiple datasets could provide more comprehensive insights into opportunities for improvement and contribute to enhancing Trauma Quality Improvement Programs on a broader scale.
-However, it is important to acknowledge that multi-center studies may compromise internal validity, reducing their applicability to specific centers.
-
-Building upon the findings of this study, future research should focus on specific variables such as ICU length of stay and the individual assessment of vital parameters like Glasgow Coma Scale, respiratory rate, and systolic blood pressure, rather than relying solely on composite scores like Revised Trauma Score.
-Analyzing these vital signs separately may reveal which parameter holds greater predictive value for opportunities for improvement.
-Also, qualitative studies could provide an in-depth understanding of why physiologically more stable patients are more likely to have opportunities for improvement.
-
-Given the significant association between higher Revised Trauma Score and increased odds of opportunities for improvement, attention should be directed towards trauma patients who appear stable upon admission but may deteriorate over time.
-Future research should aim to identify early indicators of worsening conditions in these patients, facilitating more targeted resource allocation and intervention strategies.
-
-In contrast, Injury Severity Score did not demonstrate a correlation with opportunities for improvement in this study.
-Subsequent studies could explore alternative scoring systems such as New Injury Severity Score, Revised Trauma Score, and ASA scores, or combinations of physiological and anatomical scoring systems to determine the most effective predictor of opportunities for improvement in trauma patients.
-
-Overall, future research endeavors should strive to address these gaps in knowledge to optimize trauma care outcomes and contribute to the ongoing improvement of trauma quality improvement initiatives.
+The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay longer than 7 days, and survival 30 days after injury, emphasizing the need for focused monitoring and targeted interventions for these patients.
+The results also suggest the importance of investigating specific variables like ICU length of stay and individual vital parameters, rather than composite scores such as Revised Trauma Score, to better predict opportunities for improvement.
+Future research should also explore early indicators of deterioration in seemingly stable trauma patients and evaluate alternative scoring systems to enhance predictive accuracy.
 

### 2024-09-30 manuscript2.Rmd
30/9

@@
 Notably, patients who were relatively more stable upon arrival and patients with prolonged ICU stays had the highest odds of opportunities for improvement.
-This underscores the necessity of concentrating more efforts on these specific patient groups.
+This underscores the necessity of concentrating more efforts in these  patients.
 
@@
 Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42].
-Previous research on opportunities for improvement in trauma care have been compiled into a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
+Previous research on opportunities for improvement was recently classified in a systematic review that identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].
 
@@
 
-We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital's local trauma care quality database.
-The trauma registry is part of the Swedish national trauma registry SweTrau [@76].
+We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital’s local trauma care quality database.
+The Karolinska University Hospital trauma registry is part of the Swedish national trauma registry SweTrau, and includes all patients admitted with trauma team activation, regardless of Injury Severity Score, as well as patients admitted without trauma team activation but found to have an Injury Severity Score of more than 9.
+Almost all hospitals managing patients with major trauma in Sweden contribute data to the registry [@76].
+The trauma care quality database includes data collected as part of the local peer review process.
+Screening for opportunities for improvement in selected patient cases was initiated in 2013.
+Since 2017 all patients have been included in the screening process.
 
@@
 
-The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score, as well as patients admitted without trauma team activation but found to have an Injury Severity Score of more than 9.
 We included all patients who had been included in the opportunities for improvement screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU.
@@
 
-The outcome was the presence of any opportunity for improvement.
-An opportunity for improvement presents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice.
+The outcome was the presence of any opportunity for improvement during the hospital stay, as decided by consensus during a multidisciplinary peer-review meeting.
+An opportunity for improvement represents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice.
 
@@
 The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion.
-The categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, American Society of Anesthesiologists preinjury score, and 30-day mortality.
+The categorical factors included sex [@30; @31], mechanical ventilation [@66; @78], on-call hours [@79], ICU length of stay [@22; @26], American Society of Anesthesiologists preinjury score [@33], and 30-day mortality.
 Sex was categorized as male sex or female sex.
@@
 ICU length of stay was categorized as “≤ 7 days” or “> 7 days”.
-American Society of Anesthesiologists (ASA) preinjury score was divided into “ASA 1-2” and “ASA 3-6”.
-We categorised variables to limit the number of independent variables in the multivariable analysis to adhere to the rule of thumb of ten patients with the outcome per variable [@45].
+American Society of Anesthesiologists preinjury score was divided into “ASA 1-2” and “ASA 3-5”.
+We categorized variables to limit the number of independent variables in the multivariable analysis to adhere to the rule of thumb of ten patients with the outcome per variable [@45].
 
-The continuous variables in the analysis were age in years, Revised Trauma Score, Injury Severity Score, and time to first computed tomography in minutes.
+The continuous variables in the analysis were age in years [@81], Revised Trauma Score [@11], Injury Severity Score [@77], and time to first computed tomography in minutes [@80].
 The Revised Trauma Score is a measure of severity based on prehospital or emergency department values.
-In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Come Scale scores were missing from the emergency department records, prehospital values were used instead.
+In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Coma Scale scores were missing from the emergency department records, prehospital values were used instead.
 If the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero.
@@
 Continuous variables are presented using mean values and standard deviations, and categorical variables are presented with the number of patients in each subgroup and their respective percentages.
-
-We used the Pearson’s Chi-squared test and the Wilcoxon rank sum test to compare categorical and continuous variables respectively between patients with and without opportunities for improvement.
 
@@
 A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded, as shown in Figure 1.
-Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement This left us with a total of 1,742 patients eligible for the study.
+Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement.
+This left us with a total of 1,742 patients eligible for the study.
 Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study.
@@
 
-Table 1 present sample characteristics.
+Table 1 presents sample characteristics.
 Most patients were male 1,096 (76%) vs 353 (24%).
@@
 Table 2 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.
+The most common category of opportunity for improvement in the patient cohort of this study was clinical judgement error, followed by delay in treatment.
 
@@
 Table 4 shows the results of the stratified analysis with adjusted associations between patient and process factors and opportunities for improvement among the patients who were alive and dead 30 days after injury respectively.
-The Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.
+The Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and dead `r inline_text(table3bdead, variable = RTS)` 30 days after injury respectively.
 Time to first computed tomography was significantly associated with opportunities for improvement in patients who were deceased 30 days after injury `r inline_text(table3bdead, variable = TimeFCT)`.
@@
 The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement.
-Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the unadjusted nor the adjusted analyses.
-In the stratified analysis of patients who were alive and dead 30 days after injury, only the Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
-Among the patients who deceased 30 days after injury, more than 7 days in the ICU was significantly associated with higher odds of opportunities for improvement.
+In the stratified analysis based on survival until 30 days after injury, only the Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups.
+Among the patients who were dead 30 days after injury, > 7 days in the ICU was significantly associated with higher odds of opportunities for improvement in addition to RTS.
+While these findings cannot be used to proactively identify patients at risk of medical errors or adverse events, they indicate that patients who are physiologically more stable on arrival but stay more than one week in the ICU are a potential target group for a focused review process, aiming to identify interventions to improve the quality of their care.
 
-Our study found a significant association between higher Revised Trauma Score and higher odds of opportunities for improvement, indicating that among very severely injured patients that are admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement.
-Patients with lower Revised Trauma Score are in a critical condition upon arrival — upon which emergency interventions are performed immediately.
-In contrast, patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring.
+The positive association between the Revised Trauma Score and opportunities for improvement, means that among the very severely injured patients admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement.
+One hypothetical explanation is that patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring.
 This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
@@
 The use of prehospital, emergency department, or intubation-based values for these variables may have affected data accuracy and reproducibility.
+Finally, the outcome opportunities for improvement is a consensus decision made in the local trauma quality improvement process.
+As such, these decisions have not been externally reviewed.
 
@@
 
-The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay longer than 7 days, and survival 30 days after injury, emphasizing the need for focused monitoring and targeted interventions for these patients.
-The results also suggest the importance of investigating specific variables like ICU length of stay and individual vital parameters, rather than composite scores such as Revised Trauma Score, to better predict opportunities for improvement.
-Future research should also explore early indicators of deterioration in seemingly stable trauma patients and evaluate alternative scoring systems to enhance predictive accuracy.
+The findings revealed a significant association between increased odds of opportunities for improvement and higher Revised Trauma Score, ICU length of stay longer than 7 days, and survival 30 days after injury.
+These findings indicate that patients who are physiologically more stable on arrival but stay more than one week in the ICU are a potential target group for a focused review process, aiming to identify interventions to improve the quality of their care.
 

### 2024-09-30 functions/Deceased.R

@@ -1,0 +1,16 @@
+#Använd tidigare kod för OFI för att få rätt kohort - finns i table1.R
+
+# Convert DateTime_ArrivalAtHospital and DeceasedDate to Date format
+ofi$DateTime_ArrivalAtHospital <- as.Date(ofi$DateTime_ArrivalAtHospital)
+ofi$DeceasedDate <- as.Date(ofi$DeceasedDate)
+# Calculate the difference in days
+ofi$days_to_deceased <- as.numeric(ofi$DeceasedDate - ofi$DateTime_ArrivalAtHospital)
+# Calculate the difference in days
+ofi$days_to_deceased <- as.numeric(ofi$DeceasedDate - ofi$DateTime_ArrivalAtHospital)
+# Count patients deceased within 1 day and 2 days
+deceased_within_1_day <- sum(ofi$days_to_deceased == 1, na.rm = TRUE)
+deceased_within_2_days <- sum(ofi$days_to_deceased <= 2 & ofi$days_to_deceased > 0, na.rm = TRUE)
+
+# Print the results
+cat("Patients deceased within 1 day:", deceased_within_1_day, "\n")
+cat("Patients deceased within 2 days:", deceased_within_2_days, "\n")

### 2024-09-30 functions/ofi_categories.R

@@ -1,179 +1,9 @@
-#INSTALLING PACKAGES
-#devtools::install_github("martingerdin/noacsr")
-#devtools::install_github("martingerdin/rofi")
-library(dotenv)
-library(noacsr)
-library(rofi)
-#noacsr::source_all_functions()
-data <- import_data()
-
-merged.data <- merge_data(data)
-merged.data$ofi <- create_ofi(merged.data)
-
-
-#install.packages("dplyr")
-library(dplyr)
-#install.packages("gtsummary")
-library(gtsummary)
-library(tidyverse)
-library(officer)
-
-
-#FLOWCHART: included/excluded
-#install.packages("Gmisc")
-library(Gmisc, quietly = TRUE)
-library(glue)
-#install.packages("htmlTable")
-library(htmlTable)
-library(grid)
-library(magrittr)
-
-##CLEANING DATA
-subdat <- merged.data %>%
-  select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
-         res_survival, pre_intubated, ed_intubated, dt_ed_first_ct, ISS, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
-         host_care_level, hosp_vent_days, pt_asa_preinjury, pre_gcs_sum, 
-         pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat, iva_dagar_n, Problemomrade_.FMP)
-
-#Converting subdat$ofi to logical so subset can be used 
-subdat$ofi <- ifelse(subdat$ofi == "Yes", TRUE, FALSE)
-
-#Only those in IVA
-iva <- subset(subdat, subset = (host_care_level == 5))
-
-#Removing pt_yrs < 15
-adult <- subset(iva, subset = (pt_age_yrs > 14))
-
-#Deceased on arrival 
-alive <- subset(adult, subset = (Fr1.12 == 2 | is.na(Fr1.12)))
-
-#Removing ofi = NA            
-ofi <- alive %>% subset(!is.na(ofi))
-
-
-#DEFINING VARIABLES FOR TABLE 1 
-#Gender
-ofi$Sex <- ifelse(ofi$pt_Gender == 1, "Male", 
-                  ifelse(ofi$pt_Gender == 2, "Female", 
[…]

### 2024-09-30 functions/table1.R

@@ -1,10 +1,176 @@
 
-#TABLE 1: Sample characteristics
-#Creating new table with defined data 
+#INSTALLING PACKAGES
+#devtools::install_github("martingerdin/noacsr")
+#devtools::install_github("martingerdin/rofi")
+library(dotenv)
+library(noacsr)
+library(rofi)
+#noacsr::source_all_functions()
+data <- import_data()
+
+merged.data <- merge_data(data)
+merged.data$ofi <- create_ofi(merged.data)
+
+#install.packages("dplyr")
 library(dplyr)
-library(gt)
-#install.packages("flextable")
-library(flextable)
+#install.packages("gtsummary")
+library(gtsummary)
+library(tidyverse)
+library(officer)
+
+
+#FLOWCHART: included/excluded
+#install.packages("Gmisc")
+library(Gmisc, quietly = TRUE)
+library(glue)
+#install.packages("htmlTable")
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+##CLEANING DATA
+subdat <- merged.data %>%
+  select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
+         res_survival, pre_intubated, ed_intubated, dt_ed_first_ct, ISS, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
+         host_care_level, hosp_vent_days, pt_asa_preinjury, pre_gcs_sum, 
+         pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat, iva_dagar_n, Problemomrade_.FMP)
+
+#Converting subdat$ofi to logical so subset can be used 
+subdat$ofi <- ifelse(subdat$ofi == "Yes", TRUE, FALSE)
+
+#Only those in IVA
+iva <- subset(subdat, subset = (host_care_level == 5))
+
+#Removing pt_yrs < 15
+adult <- subset(iva, subset = (pt_age_yrs > 14))
+
+#Deceased on arrival 
+alive <- subset(adult, subset = (Fr1.12 == 2 | is.na(Fr1.12)))
+
+#Removing ofi = NA            
+ofi <- alive %>% subset(!is.na(ofi))
+
+
+#DEFINING VARIABLES FOR TABLE 1 
+#Gender
[…]

### 2024-09-30 functions/table2.R

@@ -24,7 +24,7 @@ table3a <- tbl_uvregression(data = tablereg,
                               Intubation = "Mechanical ventilation",
                               ASApreinjury = "ASA preinjury",
                               OnDuty = "On call hours",
-                              TimeFCT = "Time to first CT"
+                              TimeFCT = "Time to first CT, in minutes"
                             )) %>%
   bold_labels() %>%
   bold_p(t = 0.05)
@@ -42,7 +42,7 @@ table3b <- tbl_regression(adjusted_table,
                                        Intubation = "Mechanical ventilation",
                                        OnDuty = "On call hours",
                                        ASApreinjury = "ASA preinjury",
-                                       TimeFCT = "Time to first CT")) %>%
+                                       TimeFCT = "Time to first CT, in minutes")) %>%
   bold_labels() %>%
   bold_p(t = 0.05)
 

### 2024-09-30 functions/table3.R

@@ -38,7 +38,7 @@ table3aalive <- tbl_regression(adjusted_table1,
                                        OnDuty = "On call hours",
                                        Intubation = "Mechanical ventilation",
                                        ASApreinjury = "ASA preinjury",
-                                       TimeFCT = "Time to first CT")) %>%
+                                       TimeFCT = "Time to first CT, in minutes")) %>%
   bold_labels() %>%
   add_n() %>%
   bold_p(t = 0.05)
@@ -61,7 +61,7 @@ table3bdead <- tbl_regression(adjusted_table2,
                                             OnDuty = "On call hours",
                                             Intubation = "Mechanical ventilation",
                                             ASApreinjury = "ASA preinjury",
-                                            TimeFCT = "Time to first CT")) %>%
+                                            TimeFCT = "Time to first CT, in minutes")) %>%
   bold_labels() %>%
   bold_p(t = 0.05)
 

### 2024-09-30 main.R

@@ -215,7 +215,7 @@ ofi$daysinICU <- ifelse(ofi$iva_dagar_n < 7 | ofi$iva_dagar_n == 7, "≤ 7 days"
 
 #Pt ASA preinjury
 ofi$ASApreinjury <- ifelse(ofi$pt_asa_preinjury == 1 | ofi$pt_asa_preinjury == 2, "ASA 1-2",
-                           ifelse(ofi$pt_asa_preinjury %in% 3:7, "ASA 3-7",
+                           ifelse(ofi$pt_asa_preinjury %in% 3:5, "ASA 3-5",
                                   ifelse(ofi$pt_asa_preinjury == 999, NA, NA)))
 
 #Survival after 30 days 

### 2024-11-23 manuscript2.Rmd
23/11

@@
 ### Conclusion
-Several patient and process factors were found to be associated with opportunities for improvement.
-Notably, patients who were relatively more stable upon arrival and patients with prolonged ICU stays had the highest odds of opportunities for improvement.
-This underscores the necessity of concentrating more efforts in these  patients.
+The care of patients who are physiologically more stable on admission to the ICU is more likely to be associated with opportunities for improvement.
+This association is particularly pronounced in patients who ultimately die.
+Additionally, higher odds of opportunities for improvement in survivors and those with prolonged ICU stays suggest that extended care in complex cases may both heighten error risk and lead to longer hospitalisations.
 
@@
 We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital’s local trauma care quality database.
-The Karolinska University Hospital trauma registry is part of the Swedish national trauma registry SweTrau, and includes all patients admitted with trauma team activation, regardless of Injury Severity Score, as well as patients admitted without trauma team activation but found to have an Injury Severity Score of more than 9.
-Almost all hospitals managing patients with major trauma in Sweden contribute data to the registry [@76].
+The Karolinska University Hospital trauma registry is part of the Swedish national trauma registry SweTrau which follows the Utstein template [@82].
+The registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score, as well as patients admitted without trauma team activation but found to have an Injury Severity Score of more than 9 [@76].
 The trauma care quality database includes data collected as part of the local peer review process.
@@
 The care of adult trauma patients is screened for opportunities for improvement using a combination of audit filters and individual review by specialized nurses.
-Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review meeting held every 6-8 weeks.
-The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database.
+All deceased patients, as well as those deemed as having a higher probability of opportunities for improvement, are reviewed in a multidisciplinary peer-review conducted every 6–8 weeks.
+The presence or absence of opportunities for improvement, including any corrective actions, is reached through consensus among the peer-review participants and subsequently recorded in the trauma care quality database.
+Decisions regarding the presence of preventable deaths are also made during the peer-review process..
+The different ICU:s at the Karolinska University Hospital are all Level 3 as defined by the Intensive Care Society [@83].
 
@@
 
-We included all patients who had been included in the opportunities for improvement screening process between January 1, 2014, and February 1, 2023, and who were also admitted to the ICU.
+We included all patients who had been included in the opportunities for improvement screening process between January 1, 2013, and February 28, 2023, and who were also admitted to the ICU.
 We excluded patients who were younger than 15 years and patients who were dead on arrival.
@@
 Patients who were dead on arrival were excluded as opportunities for improvement among these patients is not relevant for the aim of this study.
-
-## Study size
-
-We included all eligible patients from the trauma registry.
 
@@
 The outcome was the presence of any opportunity for improvement during the hospital stay, as decided by consensus during a multidisciplinary peer-review meeting.
-An opportunity for improvement represents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice.
+An opportunity for improvement represents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice, which are classified into the categories missed diagnosis, delay in treatment, clinical judgment errors, inadequate protocols, inadequate resources, other errors, and preventable deaths.
 
@@
 In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Coma Scale scores were missing from the emergency department records, prehospital values were used instead.
-If the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero.
-Subsequently, for the Revised Trauma Score variable, each of the variables Glasgow Coma Scale, systolic blood pressure, and respiratory rate were converted to Revised Trauma Score codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].
+If it was known that the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero when calculating the Revised Trauma Score variable.
+Subsequently, each of the variables Glasgow Coma Scale, systolic blood pressure, and respiratory rate were converted to Revised Trauma Score codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].
 
@@
 Odds ratios (OR) with associated 95% confidence intervals were calculated.
-A p-value of less than 0.05 considered indicative of a significant difference.
-As a sensitivity analysis, we considered 30-day mortality as a potential effect modifier and therefore conducted a stratified analysis in which the associations between patient and process factors and opportunities for improvement were estimated separately in patients who survived and patients who died within 30 days.
+A p-value of less than 0.05 were considered indicative of a significant difference.
+We considered 30-day mortality as a potential effect modifier and therefore conducted a stratified analysis in which the associations between patient and process factors and opportunities for improvement were estimated separately in patients who survived and patients who died within 30 days.
 All statistical analyses were conducted using R version 4.4 [@19].
@@
 Most patients were male 1,096 (76%) vs 353 (24%).
-There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than 7 days.
+Of the 354 (42%) patients who were intubated, 241 (17%) patients received mechanical ventilation for 1-7 days, while 113 (7,8%) patients received mechanical ventilation for more than 7 days.
 The number of patients arriving outside working hours was 1,042 (72%).
@@
 Amongst our patients, 30-day mortality was at 286 (20%).
+Of these, 92 patients died within 24 hours, and a total of 144 patients died within 48 hours of hospital arrival.
 When comparing patients with opportunities for improvement to those without, patients with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
@@
 Among the patients who were dead 30 days after injury, > 7 days in the ICU was significantly associated with higher odds of opportunities for improvement in addition to RTS.
-While these findings cannot be used to proactively identify patients at risk of medical errors or adverse events, they indicate that patients who are physiologically more stable on arrival but stay more than one week in the ICU are a potential target group for a focused review process, aiming to identify interventions to improve the quality of their care.
+While these findings cannot be used to proactively identify patients at risk of medical errors or adverse events, they indicate that patients who are physiologically more stable on arrival, particularly those who ultimately die, and those who stay more than one week in the ICU are potential target groups for a focused review process, aiming to identify interventions to improve the quality of their care.
 
-The positive association between the Revised Trauma Score and opportunities for improvement, means that among the very severely injured patients admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement.
-One hypothetical explanation is that patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring.
-This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
+The positive association between the Revised Trauma Score and opportunities for improvement, means that among the very severely injured patients admitted to the ICU, patients in a less severe condition on admission were at higher odds of opportunities for improvement.
+One hypothetical explanation is that patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall.
+This aligns with findings from previous research conducted on all trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
 
@@
 The ICU environment, while designed to provide advanced medical care, can also be fraught with complexities, including high patient acuity, numerous interventions, and a fast-paced workflow, all contributing to a higher cumulative exposure to potential errors the longer a patient remains in the ICU.
+
 Additionally, complications or opportunities for improvement might themselves lead to longer ICU stays.
-Conversely, shorter ICU stays may indicate patient demise before potential mistakes could occur, particularly in cases where mortality was anticipated rather than preventable.
+Conversely, shorter ICU stays may indicate patient demise before potential mistakes could occur, particularly in cases where mortality was anticipated.
+This is one potential explanation for why death was negatively associated with opportunities for improvement in our study.
 This finding is similar to previous research demonstrating prolonged ICU stays as a risk factor for increased mortality and morbidity in trauma patients [@22, @26].
@@
 
-The reason for prolonged ICU stays being associated with increased mortality has been investigated by previous studies suggesting that factors such as comorbidities, age, and illness severity upon admission do not fully elucidate the increased long-term mortality observed in this patient cohort.
+The reason for prolonged ICU stays being associated with increased mortality has been investigated by previous studies suggesting that factors such as comorbidities, age, and illness severity upon admission do not fully explain the increased long-term mortality observed in this patient cohort.
 As the duration of ICU stay increases, the primary reason for ICU stay becomes less linked to the patient’s original condition and increasingly tied to persistent chronic illness accompanied by subsequent organ dysfunction.
-The cause of the organ dysfunction comes from increased exposure to iatrogenic factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions.
-[@23, @24] Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and Injury Severity Score in our study.
-These factors primarily reflect the patient's baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and Injury Severity Score.
-
-Another factor associated with increased odds of opportunities for improvement was survival 30 days after injury.
-Previous research has identified an association between increased mortality and the severity of the injury upon admission [@59], suggesting that patients at high risk of 30-day mortality are typically in poor health upon arrival.
-Consequently, the likelihood of opportunities for improvement decreases as deaths may not be preventable.
+The cause of the organ dysfunction comes from increased exposure to iatrogenic factors and nosocomial infections, which exacerbate pre-existing organ dysfunction or lead to new-onset organ dysfunctions [@23, @24].
+Consequently, this might be a potential explanation for the nonsignificance of predictors such as ASA preinjury and Injury Severity Score in our study.
+These factors primarily reflect the patient’s baseline health status and the initial injury severity, whereas the complications that arise during prolonged ICU stays are more closely associated with the ongoing healthcare interventions and exposures experienced during hospitalization which cannot be measured with ASA preinjury and Injury Severity Score.
 

### 2024-11-23 functions/table1.R

@@ -33,7 +33,7 @@ subdat <- merged.data %>%
   select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
          res_survival, pre_intubated, ed_intubated, dt_ed_first_ct, ISS, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
          host_care_level, hosp_vent_days, pt_asa_preinjury, pre_gcs_sum, 
-         pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat, iva_dagar_n, Problemomrade_.FMP)
+         pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat, iva_dagar_n, Problemomrade_.FMP, DeceasedDate)
 
 #Converting subdat$ofi to logical so subset can be used 
 subdat$ofi <- ifelse(subdat$ofi == "Yes", TRUE, FALSE)
@@ -70,10 +70,11 @@ ofi$Intubation1 <- ifelse(ofi$pre_intubated == 1, "Intubation",
 
 #Intubation combined with ventilator days 
 ofi$Intubation <- ifelse(ofi$Intubation1 == "Not intubated", "Not intubated",
-                         ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days ==  0, "Mechnical ventilation 1-7 days",
-                                ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days %in% 1:7, "Mechanical ventilation 1-7 days",
-                                       ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days > 7, "Mechanical ventilation > 7 days", 
-                                              ifelse(ofi$Intubation1 == "Unknown", "Unknown", NA)))))
+                         ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days %in% 0:2, "Mechanical ventilation 0-2 days",
+                                ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days %in% 3:7, "Mechanical ventilation 3-7 days",
+                                       ifelse(ofi$Intubation1 == "Intubation" & ofi$hosp_vent_days > 7, "Mechanical ventilation > 7 days",
[…]

### 2024-11-23 functions/table2.R

@@ -5,7 +5,7 @@ tablereg <- ofi %>%
          ASApreinjury, Survival, OpportunityForImprovement1)
 
 tablereg$Intubation <- ifelse(is.na(tablereg$Intubation), "Unknown", table1$Intubation)
-tablereg$Intubation <- fct_relevel(tablereg$Intubation, "Not intubated", "Mechanical ventilation 1-7 days", "Mechanical ventilation > 7 days", "Unknown")
+tablereg$Intubation <- fct_relevel(tablereg$Intubation, "Not intubated", "Mechanical ventilation 0-2 days", "Mechanical ventilation 3-7 days", "Mechanical ventilation > 7 days", "Unknown")
 tablereg$daysinICU <- fct_relevel(tablereg$daysinICU, "≤ 7 days", "> 7 days")
 tablereg$Survival <- fct_relevel(tablereg$Survival, "Dead", "Alive")
 

### 2024-11-25 manuscript2.Rmd
25/11

@@
 Amongst our patients, 30-day mortality was at 286 (20%).
-Of these, 92 patients died within 24 hours, and a total of 144 patients died within 48 hours of hospital arrival.
+Of these, 105 patients died within 24 hours, and a total of 148 patients died within 48 hours of hospital arrival.
 When comparing patients with opportunities for improvement to those without, patients with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.
@@
 Table 2 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.
-The most common category of opportunity for improvement in the patient cohort of this study was clinical judgement error, followed by delay in treatment.
+The most frequently identified opportunity for improvement in the patient cohort of this study was clinical judgement errors, accounting for 42 cases (29%), followed by delays in treatment, observed in 33 cases (23%).
 
@@
 
-The positive association between the Revised Trauma Score and opportunities for improvement, means that among the very severely injured patients admitted to the ICU, patients in a less severe condition on admission were at higher odds of opportunities for improvement.
+The positive association between the Revised Trauma Score and opportunities for improvement, means that among the very severely injured patients admitted to the ICU, patients who are more physiologically stable upon admission were at higher odds of opportunities for improvement.
 One hypothetical explanation is that patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall.
-This aligns with findings from previous research conducted on all trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
+Previous research on all trauma patients at the Karolinska University Hospital Solna has demonstrated differences in opportunities for improvement between all trauma patients and those treated in the ICU [@12].
 

### 2024-11-25 functions/table1.R

@@ -164,13 +164,13 @@ ofi$Survival <- ifelse(ofi$res_survival == 1, "Dead",
 
 
 #24-hour mortality
-#ofi$DateTime_ArrivalAtHospital <- as.Date(ofi$DateTime_ArrivalAtHospital)
-#ofi$DeceasedDate <- as.Date(ofi$DeceasedDate)
+ofi$DateTime_ArrivalAtHospital <- as.Date(ofi$DateTime_ArrivalAtHospital)
+ofi$DeceasedDate <- as.Date(ofi$DeceasedDate)
 # Calculate the difference in days
-#ofi$days_to_deceased <- as.numeric(ofi$DeceasedDate - ofi$DateTime_ArrivalAtHospital)
+ofi$days_to_deceased <- as.numeric(ofi$DeceasedDate - ofi$DateTime_ArrivalAtHospital)
 # Count patients deceased within 1 day
-#ofi$Mortality <- ifelse(ofi$days_to_deceased <= 1, "Dead 24 hours", 
-#                        ifelse(ofi$days_to_deceased %in% 1:2, "Dead 48 hours", "Alive"))
+ofi$Mortality <- ifelse(ofi$days_to_deceased <= 1, "Dead",
+                        ifelse(ofi$days_to_deceased >1, "Alive", NA))
 
 #OFI 
 ofi$OpportunityForImprovement <- ifelse(ofi$ofi == TRUE, "Opportunity for improvement",
@@ -180,17 +180,19 @@ ofi$OpportunityForImprovement1 <- ifelse(ofi$OpportunityForImprovement == "Oppor
                                          ifelse(ofi$OpportunityForImprovement == "No opportunity for improvement", 0, NA))
 
 
-#table1 <- ofi %>% 
-#  select(Sex, Age, Intubation, RTS, ISS, TimeFCT, OnDuty, daysinICU, 
-#         ASApreinjury, Survival, Mortality, OpportunityForImprovement)
-
 table1 <- ofi %>% 
   select(Sex, Age, Intubation, RTS, ISS, TimeFCT, OnDuty, daysinICU, 
-         ASApreinjury, Survival, OpportunityForImprovement)
+         ASApreinjury, Survival, Mortality, OpportunityForImprovement)
+
+#table1 <- ofi %>% 
+#  select(Sex, Age, Intubation, RTS, ISS, TimeFCT, OnDuty, daysinICU, 
[…]
